---
title: "Choosing a DrugSigNet analysis strategy"
author: "DrugSigNet authors"
date: "Last update: `r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{02. Choosing a DrugSigNet analysis strategy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

DrugSigNet supports complementary signature-based, network-based, and integrated
drug-repurposing strategies. This vignette helps users choose an analysis route
based primarily on the biological evidence available for the disease.

Algorithmic details are provided in the method-specific vignettes, while
executable analyses are provided in the workflow vignettes. This vignette
therefore focuses on strategy selection, assumptions, and practical
considerations rather than implementation.

# The main decision

The first question is not which algorithm to use, but which biological input is
available and sufficiently informative for the analysis.

| Disease evidence available | Recommended strategy | Main DrugSigNet entry point |
|---|---|---|
| Signed differential-expression profile | Signature-based analysis | `drugSignaturePipeline()` or `drugRepurposingPipeline(mode = "signature")` |
| Disease-associated, risk, or seed genes | Network-based analysis | `drugNetworkPipeline()` or `drugRepurposingPipeline(mode = "network")` |
| Both a suitable expression signature and disease-gene set | Integrated analysis | `drugRepurposingPipeline(mode = "both")` |

If drug rankings have already been generated outside these workflows, users can
enter DrugSigNet at later stages such as rank aggregation, annotation, or
visualization.

# Signature-based analysis

Use signature-based analysis when the disease is represented by a signed
molecular expression profile, particularly differential-expression data.
DrugSigNet compares disease-associated expression changes with drug-induced
reference signatures using CMAP-, LINCS-, gCMAP-, and correlation-based
approaches.

Signature-based analysis is most appropriate when:

- up- and down-regulated genes or signed fold changes are available;
- the disease phenotype is meaningfully represented by transcriptomic or
  proteomic changes;
- the expression signature contains enough informative genes for comparison
  with drug-induced reference signatures; and
- the scientific question concerns perturbational relationships between disease
  and drug signatures.

A common biological interpretation is that therapeutically useful perturbations
may oppose disease-associated expression changes. However, the strength and
direction of drug-signature relationships should be interpreted in the context
of the disease state and the analysis objective.

Avoid relying exclusively on signature-based prioritization when the input
signature is weak, contains very few informative genes, or is strongly affected
by technical or compositional effects.

For methodological details, see `signature-methods.Rmd`. For an executable
analysis, see `signature-workflow.Rmd`.

# Network-based analysis

Use network-based analysis when the disease is represented by disease-associated
genes or proteins. DrugSigNet maps disease genes and drug targets onto
biological interaction networks and prioritizes drugs according to their
topological relationship to the disease module.

Network-based methods in DrugSigNet include TrustRank, harmonic centrality,
degree centrality, network proximity, and diffusion-based algorithms.

Network analysis is most appropriate when:

- the main input is a curated disease-gene or disease-protein set;
- genetic, pathway, or protein-level evidence is stronger than expression
  evidence;
- disease genes and drug targets can be mapped adequately to the interaction
  network; or
- the scientific question concerns network influence, module proximity, or
  propagation from disease seeds.

Network-based results depend on the coverage and topology of the selected
protein-protein interaction and drug-target networks. Poor mapping of disease
genes or drug targets can limit the interpretability of downstream rankings.

For methodological details, see `network-methods.Rmd`. For an executable
analysis, see `network-workflow.Rmd`.

# Integrated analysis

Integrated analysis can be considered when both a suitable signed expression
signature and a suitable disease-gene set are available.

The two evidence layers capture different properties of disease biology.
Signature-based methods evaluate relationships between disease-associated and
drug-induced molecular changes, whereas network-based methods evaluate
topological relationships between disease genes and drug targets.

Agreement between the two evidence families can strengthen confidence in a
candidate, but low agreement is not necessarily unexpected because the two
approaches capture different biological information.

In a typical integrated DrugSigNet analysis, method-specific signature and
network rankings are first summarized within their respective evidence layers
and are then combined into an integrated candidate prioritization.

For an end-to-end integrated example, see `complete-case-study.Rmd`.

# Additional considerations

Input availability alone should not determine the final strategy. Consider the
quality and suitability of each evidence source before interpreting rankings.

| Situation | Practical consideration |
|---|---|
| Weak or noisy expression signature | Avoid relying exclusively on signature-based rankings |
| Limited disease-gene or drug-target network coverage | Check mapping before interpreting network-based results |
| Uncertain disease-module compactness | Use network diagnostics before relying heavily on proximity-based methods |
| Strong disagreement among methods | Compare consensus rankings rather than selecting a single method arbitrarily |
| Both evidence types are available | Consider integrated analysis and examine agreement across evidence layers |
| Final or publication-oriented analysis | Use the strategy best supported by the available evidence and include appropriate consensus, annotation, diagnostics, and sensitivity analyses |

# Consensus ranking

DrugSigNet provides CRank, Dowdall, and Robust Rank Aggregation (RRA) to combine
method-specific rankings. Consensus ranking can help identify candidates that
receive support across multiple methods and can reduce reliance on any single
ranking algorithm.

For exploratory analyses, comparing more than one aggregation method can help
assess ranking robustness. Detailed guidance on choosing, running, and
interpreting the aggregation methods is provided in `rank-aggregation.Rmd`.

# Interpreting prioritized candidates

A high-ranking drug should be treated as a prioritization hypothesis rather
than a validated therapeutic candidate.

The strongest candidates are typically those supported by several complementary
lines of evidence, such as high ranking, consistent support across aggregation
methods, plausible targets or mechanisms, interpretable annotation profiles,
and relevant disease literature.

DrugSigNet is therefore best used to organize and integrate computational
evidence for follow-up interpretation and validation rather than as a substitute
for experimental or clinical evidence.

# Summary

This vignette described the relevant DrugSigNet concepts, inputs, outputs, and interpretation principles. Preserve the input data, function arguments, package version, and returned S4 objects when reporting or reproducing an analysis.

# Next steps

Continue with the vignette that matches the selected strategy:

| Selected task | Recommended vignette |
|---|---|
| Understand signature-based methods | `signature-methods.Rmd` |
| Run a signature-based workflow | `signature-workflow.Rmd` |
| Understand network-based methods | `network-methods.Rmd` |
| Run a network-based workflow | `network-workflow.Rmd` |
| Compare consensus methods | `rank-aggregation.Rmd` |
| Run an integrated example | `complete-case-study.Rmd` |

# Session information

Record the R and package versions used to build this vignette.

```{r session-info, eval=TRUE}
sessionInfo()
```
