---
title: "Integrated DrugSigNet workflow and complete case study"
author: "DrugSigNet authors"
date: "Last update: `r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{10. Integrated DrugSigNet workflow and complete case study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

This vignette demonstrates the integrated DrugSigNet workflow using
`drugRepurposingPipeline(mode = "both")`.

The signature and network branches are described in detail in
`signature-workflow.Rmd` and `network-workflow.Rmd`. Here, the focus is on
how those two evidence layers are run together, how their rankings are
integrated, and how to inspect the resulting signature-network consensus.

# Integrated workflow overview

With `mode = "both"`, `drugRepurposingPipeline()`:

1. runs `drugSignaturePipeline()`;
2. runs `drugNetworkPipeline()`;
3. validates that usable signature rankings are available;
4. passes the two completed pipeline objects to `integrateSignatureNetwork()`;
5. computes integrated CRank, Dowdall, and RRA rankings; and
6. returns the signature, network, and integrated sections in one
   `DrugSearchingPipeline` object.

Conceptually:

```text
Disease expression signature
          |
          v
 Signature workflow
          |
          |  Pairwise signature-family ranks
          |
          +-------------------+
                              |
                              v
                    Signature-network
                      integration
                              ^
                              |
          +-------------------+
          |
          |  Network method ranks
          |
    Network workflow
          ^
          |
   Disease seed genes
```

The integrated analysis does **not** simply combine
`Signature_CRank` with `Network_CRank`. Instead, it integrates the underlying
signature-family consensus ranks with the available network method ranks.

# Prepare inputs

The integrated workflow requires both a signed disease expression signature and
disease-associated seed genes.

```{r prepare-case-inputs}
library(DrugSigNet)

data("disease_signature", package = "DrugSigNet")
data("disease_seeds", package = "DrugSigNet")

signature_input <- disease_signature
disease_genes <- disease_seeds
```

The signature input contains `Entrez` and `FC`, while the disease-gene input is
used as the network seed set.

For analyses that use frozen signature reference databases or Synapse-backed
annotation resources, provide an authentication token securely:

```{r case-authentication}
auth_token <- Sys.getenv("SYNAPSE_AUTH_TOKEN")
```

An optional drug synonym table can be supplied through `drug_name_synonym`.
The table should be prepared before the pipeline call according to the schema
documented for `drugSignaturePipeline()`.

# Minimal integrated run

A compact first run can disable downstream annotation and visualization and use
the centrality-only network branch.

```{r minimal-integrated-run}
integrated_res <- drugRepurposingPipeline(
  signature_input = signature_input,
  disease_genes = disease_genes,
  mode = "both",
  run_all_network_methods = FALSE,
  top_k = 100,
  run_drug_annotation = FALSE,
  run_visualization = FALSE
)
```

With:

```r
run_all_network_methods = FALSE
```

the network branch runs only:

```text
TrustRank
Harmonic_centrality
Degree_centrality
```

Network proximity and diffusion are not run. Consequently, the integrated
ranking uses only those three network-derived ranks together with the
signature-family ranks.

# Reproducible case-study run

A more fully specified analysis can use frozen signature reference databases,
drug-name harmonization, and downstream annotation.

The following configuration mirrors a typical integrated analysis:

```{r integrated-case-run}
integrated_res <- drugRepurposingPipeline(
  signature_input = disease_signature,
  disease_genes = disease_seeds,
  mode = "both",
  padj = 0.05,
  drug_name_synonym = signature_drug_map,
  ties_method = "max",
  prior = 0.093,
  num_bin = 200,
  n_workers = 1,
  chunk_size = 1000,
  signature_refdb_mode = "frozen",
  signature_refdb_auth_token = auth_token,
  validate_signature_refdb = TRUE,
  top_k = 100,
  include_indirect_drugs = TRUE,
  run_all_network_methods = FALSE,
  auth_token = auth_token,
  trial_condition = "covid",
  run_drug_annotation = TRUE,
  run_visualization = TRUE
)
```

In this example, `signature_drug_map` is an analysis-specific synonym table
prepared before running the pipeline.

The explicit `padj = 0.05` is included because
`drugRepurposingPipeline()` uses `0.05` as its default signature-mode value.
This differs from the standalone `drugSignaturePipeline()` default, where
`padj = NULL`.

# Important integrated-workflow arguments

| Argument | Role in the integrated analysis |
|---|---|
| `signature_input` | Signed disease expression signature |
| `disease_genes` | Disease seed genes for the network branch |
| `mode = "both"` | Runs both branches and then integrates their rankings |
| `drug_name_synonym` | Optional drug-name harmonization for signature results |
| `ties_method` | Tie handling used during rank generation and aggregation |
| `prior` | CRank prior |
| `num_bin` | Number of CRank rank intervals |
| `chunk_size` | Signature-reference chunk size |
| `signature_refdb_mode` | Default or frozen signature reference databases |
| `signature_refdb_auth_token` | Authentication for frozen signature references |
| `validate_signature_refdb` | Validate frozen HDF5 references before use |
| `include_indirect_drugs` | Include indirectly connected drugs in the network branch |
| `run_all_network_methods` | Controls whether proximity and diffusion are added to the network branch |
| `top_k` | Threshold used for downstream annotation and visualization |
| `trial_condition` | Optional condition used for clinical-trial annotation |
| `auth_token` | Authentication used by supported Synapse-backed resources |

Signature-specific arguments are forwarded to `drugSignaturePipeline()`, while
network-specific arguments are forwarded to `drugNetworkPipeline()`.

# Inspect the combined result structure

The unified result preserves the three analysis levels:

```{r inspect-integrated-structure}
slotNames(integrated_res)

names(integrated_res@DrugSearching[["Raw"]])
names(integrated_res@DrugSearching[["Processed"]])
names(integrated_res@RankAggregation)
```

For `mode = "both"`, these sections are organized under:

```text
signature
network
integrated
```

The integrated stage itself performs rank integration rather than another drug
search. Its dedicated result is therefore primarily stored under
`@RankAggregation[["integrated"]]`.

# Inspect the signature branch

The complete signature outputs remain available in the unified object.

```{r inspect-integrated-signature}
names(
  integrated_res@DrugSearching[["Raw"]][["signature"]]
)

signature_rankings <-
  integrated_res@RankAggregation[["signature"]][["Signature_Harmonized"]]

head(signature_rankings)
```

The signature branch first calculates method-specific `rank_score`s, then
aggregates CMAP and LINCS2 reference results within CMAP, LINCS, gCMAP, and
correlation families. See `signature-workflow.Rmd` for the full processing
sequence.

# Inspect the network branch

The network outputs are stored separately:

```{r inspect-integrated-network}
names(
  integrated_res@DrugSearching[["Raw"]][["network"]]
)

network_rankings <-
  integrated_res@RankAggregation[["network"]][["Network_Harmonized"]]

head(network_rankings)
```

When `run_all_network_methods = FALSE`, `Network_Harmonized` is based on
TrustRank, harmonic centrality, and degree centrality only.

When `run_all_network_methods = TRUE`, network proximity and diffusion are also
run and incorporated into the network workflow as described in
`network-workflow.Rmd`.

# How signature and network rankings are integrated

`integrateSignatureNetwork()` first identifies drugs shared between the
signature and network results. Pipe-delimited network drug labels are expanded
before overlap matching.

Integration requires at least one overlapping drug.

## Signature contribution

For integration, DrugSigNet uses the **pairwise signature-family rankings**
rather than only the final `Signature_CRank`, `Signature_Dowdall`, and
`Signature_RRA` columns.

For CRank integration, the signature contribution is:

```text
CMAP_CRank
LINCS_CRank
gCMAP_CRank
Correlation_CRank
```

For Dowdall integration:

```text
CMAP_Dowdall
LINCS_Dowdall
gCMAP_Dowdall
Correlation_Dowdall
```

For RRA integration:

```text
CMAP_RRA
LINCS_RRA
gCMAP_RRA
Correlation_RRA
```

Each of these columns already summarizes the CMAP- and LINCS2-reference results
within one signature-search family.

## Network contribution when `run_all_network_methods = FALSE`

With the case-study setting:

```r
run_all_network_methods = FALSE
```

the network contribution is:

```text
TrustRank
Harmonic_centrality
Degree_centrality
```

The native centrality scores are converted to smaller-is-better ranks before
integration.

Therefore, the three integrated aggregation inputs are conceptually:

| Integrated ranking | Signature inputs | Network inputs |
|---|---|---|
| `Signature_Network_CRank` | `CMAP_CRank`, `LINCS_CRank`, `gCMAP_CRank`, `Correlation_CRank` | TrustRank, harmonic-centrality, degree-centrality ranks |
| `Signature_Network_Dowdall` | `CMAP_Dowdall`, `LINCS_Dowdall`, `gCMAP_Dowdall`, `Correlation_Dowdall` | TrustRank, harmonic-centrality, degree-centrality ranks |
| `Signature_Network_RRA` | `CMAP_RRA`, `LINCS_RRA`, `gCMAP_RRA`, `Correlation_RRA` | TrustRank, harmonic-centrality, degree-centrality ranks |

This preserves the individual signature-method families and network methods as
separate evidence sources during integration.

## If `run_all_network_methods = TRUE`

When the full network workflow is enabled, the integration step can additionally
include:

```text
Network_proximity
DSD-min-Rank
KL-med-Rank
KL-min-Rank
JS-med-Rank
JS-min-Rank
```

Thus, enabling all network methods changes not only the network consensus but
also the network evidence supplied to the integrated rank aggregation.

# Integrated rank aggregation

All inputs supplied to the integration step are rank-oriented, with smaller
values representing stronger priority.

DrugSigNet therefore uses:

```text
CRank:   reverse = TRUE
Dowdall: reverse = FALSE
RRA:     reverse = FALSE
```

The three integrated aggregation objects are available as:

```{r inspect-integrated-aggregation}
integrated_rank_aggregation <-
  integrated_res@RankAggregation[["integrated"]]

names(integrated_rank_aggregation)

integrated_rank_aggregation[["Signature_Network_CRank"]]@result
integrated_rank_aggregation[["Signature_Network_Dowdall"]]@result
integrated_rank_aggregation[["Signature_Network_RRA"]]@result
```

Their final rank columns are:

```text
Signature_Network_CRank
Signature_Network_Dowdall
Signature_Network_RRA
```

Smaller values indicate higher integrated priority.

# Final integrated ranking table

The analysis-ready integrated table is:

```{r inspect-integrated-harmonized}
integrated_rankings <-
  integrated_res@RankAggregation[["integrated"]][[
    "Signature_Network_Harmonized"
  ]]

head(integrated_rankings)
colnames(integrated_rankings)
```

`Signature_Network_Harmonized` retains the overlapping signature and network
ranking information together with the three new integrated consensus ranks.

The final integrated ranks can be inspected directly:

```{r top-integrated-candidates}
integrated_rankings |>
  dplyr::arrange(Signature_Network_CRank) |>
  dplyr::select(
    Drug,
    Signature_Network_CRank,
    Signature_Network_Dowdall,
    Signature_Network_RRA
  ) |>
  head(20)
```

The three aggregation methods should be interpreted as complementary consensus
formulations rather than assuming that one is universally preferable.

# Compare evidence layers

Because the unified object retains the signature, network, and integrated
tables, candidate support can be compared across analysis layers without
rerunning the pipelines.

```{r compare-analysis-layers}
signature_top <- signature_rankings |>
  dplyr::arrange(Signature_CRank) |>
  dplyr::select(
    Drug,
    Signature_CRank,
    Signature_Dowdall,
    Signature_RRA
  )

network_top <- network_rankings |>
  dplyr::arrange(Network_CRank) |>
  dplyr::select(
    Drug,
    Network_CRank,
    Network_Dowdall,
    Network_RRA
  )

integrated_top <- integrated_rankings |>
  dplyr::arrange(Signature_Network_CRank) |>
  dplyr::select(
    Drug,
    Signature_Network_CRank,
    Signature_Network_Dowdall,
    Signature_Network_RRA
  )
```

A high integrated rank can therefore be interpreted alongside the underlying
signature and network evidence rather than as an isolated score.

# Annotation and enrichment

When `run_drug_annotation = TRUE`, annotation sections from all stages are
retained in the unified object.

```{r inspect-integrated-annotation}
names(integrated_res@DrugAnnotation)

integrated_annotation <-
  integrated_res@DrugAnnotation[["integrated"]]

integrated_annotation
```

The integrated annotation section can contain:

```text
top_k_union_drugs
Features
Functional_Enrichment
```

Detailed drug annotation and `TSEA()` usage are covered in
`drug-annotation.Rmd`.

# Visualization

When `run_visualization = TRUE`, visualization outputs are similarly organized
by analysis stage:

```{r inspect-integrated-visualization}
names(integrated_res@Visualization)

integrated_res@Visualization[["integrated"]]
```

For standalone plotting of the integrated ranking table, use the current
`plot_*()` functions documented in `visualization.Rmd`.

For example, rank agreement among the three integrated consensus rankings can
be inspected with:

```{r integrated-rank-agreement}
agreement <- plot_rank_agreement(
  drug_ranks_df = integrated_rankings |>
    dplyr::select(
      Drug,
      Signature_Network_CRank,
      Signature_Network_Dowdall,
      Signature_Network_RRA
    ),
  cor_method = "spearman",
  top_k = 100
)

agreement$plot
```

# Candidate interpretation

Integrated prioritization is most useful when the final rank is interpreted
together with its contributing evidence.

For each high-priority drug, consider:

- whether it is supported across several signature-method families;
- whether it is supported by more than one network method;
- whether the signature and network evidence are concordant or complementary;
- the direction of the underlying signature association;
- plausible drug targets and mechanisms;
- known indications, development status, and safety information;
- condition-relevant clinical evidence; and
- whether the candidate provides a biologically meaningful hypothesis for
  follow-up.

A high integrated rank indicates convergence within the computational framework;
it does not by itself establish therapeutic efficacy.

# Save, export, and report results

The complete object retains the signature, network, and integrated results and
can be saved for downstream analysis.

```{r save-integrated-result}
saveRDS(
  integrated_res,
  file = "drugsignet_integrated_case_study.rds"
)
```

A compact ranking table can also be exported directly:

```{r save-integrated-table}
write.csv(
  integrated_rankings,
  file = "drugsignet_integrated_rankings.csv",
  row.names = FALSE
)
```

## Export pipeline results to Excel

`write_pipeline_results()` writes available pipeline tables to a multi-sheet
Excel workbook. Depending on the result object, the workbook can include raw
drug-search results, harmonized rankings, drug annotations, top-ranked drugs,
and functional enrichment results.

```{r export-integrated-results}
write_pipeline_results(
  result_obj = integrated_res,
  file_path = "DrugSigNet_integrated_results.xlsx",
  top_n = 100
)
```

If `file_path` is a directory, the default workbook name
`drugsignet_pipeline_results.xlsx` is used inside that directory.

## Generate an analysis report

`write_report()` generates an HTML or PDF summary of one or more DrugSigNet
pipeline results. Reports can include drug-search results, rank aggregation,
annotation, and visualization sections when available.

```{r write-integrated-report}
write_report(
  object = integrated_res,
  title = "DrugSigNet integrated analysis",
  file = "DrugSigNet_integrated_report",
  device = "html",
  interactive = TRUE
)
```

Set `write_results = TRUE` to export the corresponding Excel result workbook
while generating the report:

```{r write-integrated-report-with-results}
write_report(
  object = integrated_res,
  file = "DrugSigNet_integrated_report",
  device = "html",
  interactive = TRUE,
  write_results = TRUE
)
```

Use `device = "pdf"` for PDF output. PDF rendering requires an available LaTeX
engine. HTML reports use the tabset layout by default; set
`dashboard = "shinydashboard"` when an interactive Shiny dashboard is preferred.

# Reproducibility checklist

For a manuscript or shared analysis, report:

- DrugSigNet version;
- disease-signature source and preprocessing;
- disease-gene source and preprocessing;
- identifier systems;
- CMAP/LINCS2 reference resources and whether frozen references were used;
- PPI and drug-target network sources;
- whether `run_all_network_methods` was `TRUE` or `FALSE`;
- `ties_method`, `prior`, and `num_bin`;
- relevant signature and network parameters;
- `top_k`;
- annotation sources and trial condition when used; and
- versions or retrieval dates of external resources.

# Troubleshooting

## No overlapping drugs are available for integration

The integration step requires drugs shared between the signature pairwise
rankings and the network ranking table. Check drug-name harmonization and the
drug universes represented by the two workflows.

## Signature and network results are present but integrated results are absent

Confirm that the signature result contains usable
`RankAggregation[["Pairwise"]]` CRank, Dowdall, and RRA results and that the
network result contains usable network-method rankings.

## Integrated results change when `run_all_network_methods` changes

This is expected. With `FALSE`, integration uses the three centrality-derived
network ranks. With `TRUE`, network-proximity and diffusion ranks can also enter
the integrated aggregation.

## Frozen reference loading fails

Check `signature_refdb_auth_token`, `auth_token`, Synapse access, local cache
permissions, and HDF5 validation.

# Summary

This vignette described the relevant DrugSigNet concepts, inputs, outputs, and interpretation principles. Preserve the input data, function arguments, package version, and returned S4 objects when reporting or reproducing an analysis.

# Next steps

| Goal | Recommended vignette |
|---|---|
| Review individual signature methods | `signature-methods.Rmd` |
| Review individual network methods | `network-methods.Rmd` |
| Understand CRank, Dowdall, and RRA | `rank-aggregation.Rmd` |
| Annotate integrated candidates | `drug-annotation.Rmd` |
| Visualize integrated rankings | `visualization.Rmd` |
| Review the signature branch in detail | `signature-workflow.Rmd` |
| Review the network branch in detail | `network-workflow.Rmd` |

# Session information

Record the R and package versions used to build this vignette.

```{r session-info, eval=TRUE}
sessionInfo()
```
