---
title: "Drug annotation with DrugSigNet"
author: "DrugSigNet authors"
date: "Last update: `r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{06. Drug annotation with DrugSigNet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

Drug annotation converts prioritized drug names into interpretable biological,
chemical, safety, regulatory, and clinical information.

DrugSigNet provides two complementary annotation interfaces:

- `annotate_drugs()` retrieves and combines multiple annotation layers in one
  data frame; and
- individual `get_drug_*()` functions retrieve a specific annotation layer and
  return a `DrugAnnotation` object containing both results and parameter
  provenance.

DrugSigNet also provides `TSEA()` for target set enrichment. Unlike the drug
annotation functions, `TSEA()` operates on a **gene target list**, not directly
on drug names.

# What `annotate_drugs()` can retrieve

`annotate_drugs()` combines available annotation layers for one or more drugs,
including:

- drug synonyms;
- mechanisms of action;
- canonical SMILES;
- blood-brain barrier permeability;
- Anatomical Therapeutic Chemical (ATC) classifications;
- therapeutic indications;
- regulatory approval or development status;
- adverse-event, warning, or toxicity information; and
- clinical trial information when a `condition` is supplied.

The function returns a data frame with the available annotations merged by drug.
Annotation layers that cannot be resolved for a requested source are handled
internally so that one unavailable layer does not necessarily prevent retrieval
of the remaining annotations.

# Authentication and cached annotation resources

> **Synapse authentication**
>
> Some annotation resources require a Synapse Personal Access Token. Configure
> authentication before running these functions; see the
> [Getting Started vignette](getting-started.Rmd).

Several DrugSigNet annotation resources are loaded from a local cache or
Synapse. The configured token can be read from the environment when it is
needed:

```{r annotation-authentication}
auth_token <- Sys.getenv("SYNAPSE_AUTH_TOKEN")
```

The `force = TRUE` argument can be used by annotation helpers to refresh
supported cached resources instead of reusing an existing local copy.

# Combined annotation with `annotate_drugs()`

For broad annotation coverage, use `annotate_drugs()`:

```{r annotate-drugs}
drugs <- c(
  "metformin",
  "nifedipine",
  "cyclosporine",
  "niclosamide"
)

ann <- annotate_drugs(
  drugs = drugs,
  source = "All",
  auth_token = auth_token
)

head(ann)
```

`drugs` can be supplied as a character vector, a one-column data frame, or a
data frame containing a `Drug` column.

The supported top-level sources are:

```text
"All"
"CHEMBL"
"OpenTargets"
"TTD"
```

Use `source = "All"` when broad coverage across available annotation resources is
preferred.

## Include condition-specific clinical trials

Clinical trial information is added by `annotate_drugs()` only when `condition`
is supplied.

```{r annotate-drugs-with-trials}
ann_trials <- annotate_drugs(
  drugs = c("Imatinib", "Gefitinib"),
  source = "All",
  condition = "cancer",
  auth_token = auth_token
)

head(ann_trials)
```

When matching trials are available, the merged annotation can include clinical
trial conditions, phases, and NCT identifiers.

# Individual annotation functions

Use the individual helpers when only one annotation layer is required or when
the annotation source needs to be controlled more precisely.

| Function | Annotation retrieved | Supported source choices |
|---|---|---|
| `get_drug_synonyms()` | Alternative drug names and synonyms | `"All"`, `"OpenTargets"`, `"CHEMBL"`, `"TTD"` |
| `get_drug_status()` | Regulatory approval or development status | `"All"`, `"OpenTargets"`, `"CHEMBL"`, `"TTD"` |
| `get_drug_indications()` | Therapeutic indications | `"All"`, `"OpenTargets"`, `"CHEMBL"`, `"TTD"` |
| `get_drug_moa()` | Mechanisms of action | `"All"`, `"OpenTargets"`, `"CHEMBL"` |
| `get_drug_bbb()` | Blood-brain barrier permeability | Curated DrugSigNet BBB annotation |
| `get_drug_atc()` | ATC classifications | `"All"`, `"WHO"`, `"CHEMBL"` |
| `get_drug_adverse_events()` | Adverse-event, warning, and toxicity information | `"All"`, `"OpenTargets"`, `"CHEMBL"` |
| `get_drug_smiles()` | Canonical SMILES strings | `"All"`, `"OpenTargets"`, `"CHEMBL"`, `"TTD"`, `"B3DB_classification"` |
| `get_drug_trials()` | Clinical trial records | DrugSigNet clinical-trial resource; optional condition filter |

The source choices are layer-specific. A source accepted by one helper is not
necessarily available for every annotation type.

# Run individual annotation methods

The examples below use the same small drug set to demonstrate the individual
interfaces.

```{r individual-drug-input}
drug_query <- c("Imatinib", "Gefitinib")
```

## Drug synonyms

```{r get-drug-synonyms}
synonym_res <- get_drug_synonyms(
  drugs = drug_query,
  source = "CHEMBL",
  auth_token = auth_token
)

synonym_res@result
```

## Drug status

```{r get-drug-status}
status_res <- get_drug_status(
  drugs = drug_query,
  source = "All",
  auth_token = auth_token
)

status_res@result
```

## Drug indications

```{r get-drug-indications}
indication_res <- get_drug_indications(
  drugs = drug_query,
  source = "All",
  auth_token = auth_token
)

indication_res@result
```

## Mechanism of action

```{r get-drug-moa}
moa_res <- get_drug_moa(
  drugs = drug_query,
  source = "CHEMBL",
  auth_token = auth_token
)

moa_res@result
```

## Blood-brain barrier permeability

```{r get-drug-bbb}
bbb_res <- get_drug_bbb(
  drugs = drug_query,
  auth_token = auth_token
)

bbb_res@result
```

## ATC classification

```{r get-drug-atc}
atc_res <- get_drug_atc(
  drugs = drug_query,
  source = "CHEMBL",
  auth_token = auth_token
)

atc_res@result
```

## Adverse events and toxicity

```{r get-drug-adverse-events}
adverse_res <- get_drug_adverse_events(
  drugs = drug_query,
  source = "All",
  auth_token = auth_token
)

adverse_res@result
```

## SMILES

```{r get-drug-smiles}
smiles_res <- get_drug_smiles(
  drugs = drug_query,
  source = "CHEMBL",
  auth_token = auth_token
)

smiles_res@result
```

The annotation helpers also accept a one-column data frame when supported by the
individual function:

```{r individual-drug-data-frame}
drug_df <- data.frame(
  drug = c("Cetirizine", "Pentazocine")
)

smiles_df <- get_drug_smiles(
  drugs = drug_df,
  source = "All",
  auth_token = auth_token
)

smiles_df@result
```

## Clinical trials

`get_drug_trials()` retrieves clinical trial records in which the supplied drug
names match the intervention field.

```{r get-drug-trials}
trial_res <- get_drug_trials(
  drugs = drug_query,
  auth_token = auth_token
)

trial_res@result
```

An optional disease or condition can be used to restrict the clinical trial
records before drug matching:

```{r get-drug-trials-condition}
trial_condition_res <- get_drug_trials(
  drugs = drug_query,
  condition = "cancer",
  ignore_case = TRUE,
  auth_token = auth_token
)

trial_condition_res@result
```

`ignore_case = TRUE` performs case-insensitive matching. The returned trial
records include a `matched_drug` field indicating which supplied drug matched
the intervention record.

# Inspecting results and parameters

The individual `get_drug_*()` functions return `DrugAnnotation` S4 objects.
Their annotation table is stored in `@result`, while `@parameters` retains the
input and settings used for the retrieval.

For example:

```{r inspect-annotation-parameters}
moa_res@result
names(moa_res@parameters)

# Typical fields:
#> [1] "input_data" "source" "force" "auth_token"
```

The exact parameter fields depend on the annotation function. For example,
`get_drug_trials()` additionally stores the condition and case-matching option.

```{r inspect-trial-parameters}
names(trial_condition_res@parameters)

# Includes:
# input_data, condition, ignore_case, force, auth_token
```

Keeping `@parameters` with an individual annotation result makes it possible to
recover which drugs, source, and retrieval settings were used.

By contrast, `annotate_drugs()` is a convenience function that merges available
layers and returns the combined annotation directly as a data frame.

# Choosing annotation sources

Different resources contribute different kinds of evidence.

| Source | Typical role in DrugSigNet annotation |
|---|---|
| `"All"` | Broad annotation coverage across available sources |
| `"OpenTargets"` | Drug, indication, mechanism, status, and supported safety information |
| `"CHEMBL"` | Drug, mechanism, chemical structure, ATC, indication, status, and supported safety information |
| `"TTD"` | Drug synonyms, indications, status, and supported SMILES information |
| `"WHO"` | ATC classification through `get_drug_atc()` |
| `"B3DB_classification"` | SMILES information available to `get_drug_smiles()` |
| DrugSigNet BBB resource | Curated blood-brain barrier permeability annotations |
| DrugSigNet clinical-trial resource | Clinical trial records, optionally filtered by condition |

No single source is expected to contain every annotation for every compound.
Missing annotation should therefore be treated as missing source coverage rather
than evidence that a property is absent.

# Target set enrichment with `TSEA()`

Target set enrichment asks whether the molecular targets represented by a set of
prioritized drugs converge on particular biological functions.

Importantly, `TSEA()` receives **target genes**, not drug names.

The current interface is:

```r
TSEA(
  object = NULL,
  targetList,
  source = "GO",
  ont = c("BP", "MF", "CC", "ALL"),
  Adjusted.P.value = 0.05,
  target_id_from = NULL
)
```

## Gene Ontology enrichment

By default, targets are treated as HGNC gene symbols unless another identifier
type is supplied or Ensembl gene identifiers are detected automatically.

```{r target-enrichment}
target_genes <- c("TP53", "BRCA1", "EGFR")

tsea <- TSEA(
  targetList = target_genes,
  source = "GO",
  ont = "BP",
  Adjusted.P.value = 0.05
)

tsea@result
```

For Gene Ontology enrichment, `ont` can be:

- `"BP"`: Biological Process;
- `"MF"`: Molecular Function;
- `"CC"`: Cellular Component; or
- `"ALL"`: all three GO namespaces.

The current DrugSigNet enrichment backend uses the GO databases selected by
`ont`; this vignette therefore documents `source = "GO"` for the implemented
target-set enrichment workflow.

## Non-symbol gene identifiers

When targets are not HGNC symbols, `target_id_from` can specify the source
identifier type. Supported aliases include Ensembl, Entrez, and symbol-based
identifiers.

```{r target-enrichment-ensembl}
ensembl_targets <- c(
  "ENSG00000141510",
  "ENSG00000012048"
)

tsea_ensembl <- TSEA(
  targetList = ensembl_targets,
  source = "GO",
  ont = "ALL",
  target_id_from = "ENSEMBL",
  Adjusted.P.value = 0.05
)

tsea_ensembl@result
```

When `target_id_from = NULL`, DrugSigNet attempts to detect Ensembl gene IDs;
otherwise targets are treated as symbols.

## Inspect enrichment provenance

`TSEA()` also returns a `DrugAnnotation` object, so the enrichment input and
settings can be retained with the result:

```{r inspect-tsea}
tsea@result
tsea@parameters
```

The result is filtered using `Adjusted.P.value`.

Drug annotation can also be performed automatically within the main DrugSigNet
pipelines by setting `run_drug_annotation = TRUE`. Pipeline-generated annotation
results are stored in the `DrugAnnotation` slot. See the signature, network, and
integrated workflow vignettes for complete examples.

This avoids rerunning a separate annotation workflow when annotation was already
requested during the main analysis.

# Interpreting annotation

Annotation should be used to contextualize and prioritize computational
candidates rather than to establish therapeutic efficacy.

Useful evidence may include:

- reproducible ranking support across methods;
- a plausible mechanism of action;
- relevant therapeutic indications;
- regulatory or development status;
- blood-brain barrier properties when relevant to the disease context;
- interpretable safety or toxicity annotations;
- disease-relevant clinical trial records;
- chemical structure information; and
- convergence of drug targets on relevant biological processes.

Clinical trial presence alone does not imply efficacy. Trial condition, phase,
status, intervention role, and available outcomes should be interpreted
carefully.

Likewise, absence of an annotation does not establish absence of the biological
property; it may simply reflect incomplete source coverage.

# Practical candidate table

A practical candidate table can combine ranking evidence with annotation and
manual interpretation.

| Column | Description |
|---|---|
| `Drug` | Harmonized drug name |
| Best rank | Best or selected consensus rank |
| Consensus support | Support across CRank, Dowdall, RRA, or analysis families |
| Mechanism of action | Retrieved mechanism annotation |
| Indications | Known therapeutic indications |
| Approval/development status | Regulatory or development annotation |
| ATC | Therapeutic classification |
| BBB | Blood-brain barrier annotation when available |
| Safety | Adverse-event, warning, or toxicity annotation |
| Clinical trials | Relevant trial condition, phase, or NCT identifiers |
| Targets | Drug targets obtained from the drug-target network or other target source |
| Target enrichment | Relevant TSEA terms for the selected target set |
| Interpretation | Brief biological/manual rationale |

The drug-target list used for `TSEA()` is conceptually separate from the
drug-level annotation returned by `annotate_drugs()`.

# Save annotation output

The merged data frame or individual S4 annotation objects can be saved for
reproducible downstream analysis.

```{r save-annotation}
saveRDS(ann, file = "inst/extdata/example_drug_annotation.rds")
saveRDS(moa_res, file = "inst/extdata/example_moa_annotation.rds")
saveRDS(tsea, file = "inst/extdata/example_tsea.rds")
```

# Common pitfalls

## Passing drugs directly to `TSEA()`

`TSEA()` expects gene targets in `targetList`. Obtain or construct the target
set first, then run enrichment.

## Drug names do not map

Drug matching depends on the available annotation names and synonyms. Use
`get_drug_synonyms()` to inspect alternative names when expected annotations are
missing.

## Assuming every source supports every annotation layer

Annotation-source support differs by helper. Use the source choices documented
for the individual function rather than assuming that all sources are
interchangeable.

## Annotation is incomplete

No single resource covers every compound. Report missing annotation fields
transparently and avoid treating missing values as negative biological evidence.

## Clinical evidence is ambiguous

A trial involving a drug is evidence that the drug has been investigated in that
clinical context, not evidence that the intervention was effective.

## Cached resources are stale or unavailable

Use `force = TRUE` when a supported resource should be refreshed and ensure that
the required Synapse authentication is available when the requested dataset is
not already cached locally.

# Summary

This vignette described the relevant DrugSigNet concepts, inputs, outputs, and interpretation principles. Preserve the input data, function arguments, package version, and returned S4 objects when reporting or reproducing an analysis.

# Next steps

| Goal | Recommended vignette |
|---|---|
| Understand how candidates were ranked | `rank-aggregation.Rmd` |
| Visualize ranking and annotation results | `visualization.Rmd` |
| Run a complete signature analysis | `signature-workflow.Rmd` |
| Run a complete network analysis | `network-workflow.Rmd` |
| Run integrated prioritization and interpretation | `complete-case-study.Rmd` |

# Session information

Record the R and package versions used to build this vignette.

```{r session-info, eval=TRUE}
sessionInfo()
```
