---
title: "Network-based methods in DrugSigNet"
author: "DrugSigNet authors"
date: "Last update: `r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{04. Network-based methods in DrugSigNet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

This vignette explains the network-based drug-prioritization methods implemented
in DrugSigNet: TrustRank, harmonic centrality, degree centrality, network
proximity, and diffusion-based scoring.

The focus is on what each method measures, the network inputs it uses, the
result columns it returns, and how the resulting scores should be interpreted.
For an end-to-end executable workflow, see `network-workflow.Rmd`.

# Biological rationale

Network-based drug repurposing assumes that disease-associated genes and drug
targets occupy informative positions in a biological interaction network. A
drug may be relevant when its targets are strongly connected with, close to, or
topologically similar to the disease module.

DrugSigNet uses complementary network formulations because they capture
different relationships:

- **TrustRank** measures disease-seeded random-walk influence;
- **harmonic centrality** measures weighted shortest-path accessibility from
  disease seeds;
- **degree centrality** uses a penalized multi-hop neighborhood score around
  disease seeds;
- **network proximity** measures drug-target distance to the disease module
  relative to a degree-matched random background; and
- **diffusion methods** compare propagated network representations using
  diffusion-state distance and divergence measures.

These methods should therefore be interpreted as complementary rather than as
different numerical estimates of the same quantity.

# Network inputs

All network methods use disease-associated genes together with a
protein-protein interaction network and a drug-target interaction network.

| Input | Expected structure | Role |
|---|---|---|
| `disease_genes` | Data frame with a `gene` column | Defines disease seed genes |
| `ppi_network` | Data frame with `gene1` and `gene2` columns | Defines protein-protein interactions |
| `drug_target_network` | Data frame with `ID`, `Drug`, `Target`, and `Group` columns | Maps drugs to molecular targets |

If `ppi_network` or `drug_target_network` is `NULL`, DrugSigNet can resolve the
corresponding default network with `load_drugsignet_network()` from the local
cache or Synapse.

Gene identifiers in `disease_genes`, `ppi_network`, and `drug_target_network`
must be compatible. Network results should not be interpreted before checking
how many disease seeds and drug targets map to the selected interactome.

# Inspecting results and parameters

Each individual network method returns a `NetworkBased` object with two
important slots:

| Slot | Contents |
|---|---|
| `@result` | Method-specific drug prioritization results |
| `@parameters` | Input networks, disease genes, method name, and retained method-specific settings |

For example:

```{r inspect-network-parameters}
names(trust_rank_result@parameters)
```

For a TrustRank search, `@parameters` retains the resolved `ppi_network`,
`drug_target_network`, and `disease_genes` together with settings such as
`target`, drug-inclusion flags, `filter_paths`, `hub_penalty`,
`damping_factor`, `max_deg`, and `result_size`.

The exact parameter fields differ by method. DrugSigNet filters the parameter
slot so that each method retains the inputs and settings relevant to that
specific analysis.

Important method-specific settings include:

| Method | Examples of retained settings |
|---|---|
| TrustRank | `hub_penalty`, `damping_factor`, `max_deg`, `result_size` |
| Harmonic centrality | `hub_penalty`, `max_deg`, `result_size` |
| Degree centrality | `max_deg`, `result_size` |
| Network proximity | `n_simulations`, worker/range settings, `random_seed` |
| Diffusion | `ties_method`, `output_dir` |

Because `@parameters` also retains the actual network and disease-gene inputs,
it is useful for reproducing or auditing a particular network search. The
individual examples below therefore focus on `@result`; the parameter slot only
needs to be inspected once when the provenance or settings of a run are needed.

# Centrality-based methods

TrustRank, harmonic centrality, and degree centrality return a two-column result:

```text
Drug    Score
```

For all three methods, **higher `Score` values indicate stronger network-based
prioritization**. Their scores should not, however, be compared numerically
across methods because they are generated by different algorithms.

## TrustRank

### How the method works

TrustRank is a disease-seeded PageRank-like propagation method. Disease genes
define a personalized trust vector, and relevance is propagated over the
integrated protein-protein/drug-target graph.

DrugSigNet applies hub-aware edge weighting and uses a damping factor to control
the balance between continued propagation through the network and return to the
disease-seed distribution.

A drug receives a high score when its node accumulates substantial
disease-seeded random-walk relevance.

### Main parameters

| Parameter | Interpretation |
|---|---|
| `damping_factor` | Controls the persistence of the random walk; default `0.95` |
| `hub_penalty` | Penalizes paths through highly connected hubs; default `0.01` |
| `max_deg` | Excludes nodes above a maximum degree threshold when requested |
| `result_size` | Number of top candidates returned |
| `include_indirect_drugs` | Controls inclusion of drugs not directly connected to disease genes |
| `include_non_approved_drugs` | Controls inclusion of non-approved/investigational drugs |

### Result interpretation

`TrustRank()` returns:

| Column | Meaning |
|---|---|
| `Drug` | Drug name |
| `Score` | Disease-seeded TrustRank score |

Higher `Score` values correspond to higher priority.

### Main considerations

TrustRank is disease-specific because the propagation starts from the supplied
seed genes. Results therefore depend on seed quality, interactome coverage,
hub handling, and the damping factor.

## Harmonic centrality

### How the method works

Harmonic centrality summarizes accessibility from disease seeds through inverse
shortest-path distance. In the current DrugSigNet implementation, weighted
shortest-path distances are calculated from each mapped disease seed and their
inverse contributions are accumulated for candidate nodes.

Unlike classical closeness centrality, harmonic centrality remains well defined
when parts of the network are disconnected because unreachable nodes contribute
no useful inverse-distance signal rather than making the statistic undefined.

### Main parameters

The main method-specific setting is `hub_penalty`, which affects the edge
weights used for shortest-path calculations. `max_deg` and `result_size` can
also restrict the candidate search.

### Result interpretation

`Harmonic_centrality()` returns:

| Column | Meaning |
|---|---|
| `Drug` | Drug name |
| `Score` | Accumulated harmonic accessibility from disease seeds |

Higher `Score` values correspond to stronger accessibility from the disease
seed set and therefore higher priority.

### Main considerations

The method is useful for distributed disease modules and disconnected
interactomes, but shortest-path structure may not fully represent biological
signal flow and broadly central nodes can still receive high scores.

## Degree centrality

### How the method works

Despite the function name, the current DrugSigNet `Degree_centrality()`
implementation is not simply the raw graph degree of each drug node.

For every disease seed, non-seed nodes receive:

- `1` for membership in the 1-hop neighborhood; and
- `0.5` for membership in the 2-hop neighborhood, excluding nodes already
  counted at 1 hop.

These contributions are accumulated across disease seeds. The resulting score
therefore represents a penalized local neighborhood relationship to the disease
seed set.

### Result interpretation

`Degree_centrality()` returns:

| Column | Meaning |
|---|---|
| `Drug` | Drug name |
| `Score` | Accumulated 1-hop/2-hop disease-neighborhood score |

Higher `Score` values indicate stronger local connectivity with the disease
seed neighborhood and therefore higher priority.

### Main considerations

This method is computationally simple and provides a useful local-topology
baseline. Because it emphasizes short paths around disease seeds, results can
still be influenced by dense network regions and should be interpreted
alongside more global methods.

# Network proximity

## How the method works

Network proximity quantifies how close a drug's mapped targets are to the
disease module in the largest connected component of the interactome.

For each drug target, the method identifies its shortest distance to the
disease-gene set and averages these closest distances across the drug targets.
The observed distance is compared with a degree-matched randomized background
generated through repeated simulations.

The standardized proximity score is:

\[
Z = \frac{D - \mathrm{Mean}}{\mathrm{SD}},
\]

where `D` is the observed drug-target/disease distance and `Mean` and `SD`
describe the randomized null distribution.

## Result columns

`Network_proximity()` returns:

```text
Drug    Disease    D    Z    Mean    SD
```

| Column | Meaning |
|---|---|
| `Drug` | Drug name |
| `Disease` | Disease-set label used by the proximity backend |
| `D` | Observed mean closest shortest-path distance between drug targets and disease genes |
| `Z` | Standardized proximity relative to the randomized null |
| `Mean` | Mean proximity from the degree-matched random simulations |
| `SD` | Standard deviation of the randomized proximity distribution |

For interpretation:

- **smaller `D`** means the drug targets are closer to the disease genes in raw
  network distance;
- **more negative `Z`** means the observed drug-target proximity is closer than
  expected under the degree-matched random background;
- `Z` near zero indicates proximity similar to the null expectation; and
- positive `Z` indicates greater separation than expected under the null.

Because the standardized score accounts for network-degree structure, `Z` is
generally more informative for proximity interpretation than `D` alone.

## Main parameters

| Parameter | Interpretation |
|---|---|
| `n_simulations` | Number of random simulations used to estimate the null distribution |
| `n_workers` | Number of workers used for the simulation procedure |
| `random_seed` | Seed for reproducible randomization |

## Main considerations

Network proximity is most informative when disease genes and drug targets map
adequately to the interactome and the disease genes define a meaningful network
module. Dense interactomes, very large seed sets, fragmented disease modules,
or drugs with very few mapped targets can reduce discriminatory power.

# Diffusion-based methods

## How the method works

DrugSigNet's diffusion workflow constructs network-based similarity/distance
representations and summarizes the relationship between drug targets and disease
seeds using:

- **DSD**: diffusion-state distance;
- **KL**: Kullback-Leibler divergence; and
- **JS**: Jensen-Shannon divergence.

The implementation generates or reuses cached diffusion matrices for the PPI
network. For each drug, mapped drug targets are compared with the disease seeds,
and distance/divergence summaries are converted to ranks.

## Result columns

The diffusion result contains:

```text
Drug
APP-Drugs
Degree
Targets
DSD-min-Rank
KL-med-Rank
KL-min-Rank
JS-med-Rank
JS-min-Rank
DSD-min
KL-med
KL-min
JS-med
JS-min
```

The columns can be grouped as follows:

| Column(s) | Meaning |
|---|---|
| `Drug` | Drug name |
| `APP-Drugs` | Backend drug-status indicator retained in the diffusion output |
| `Degree` | Number of mapped targets used for the drug |
| `Targets` | Set of mapped drug targets represented in the network |
| `DSD-min`, `KL-med`, `KL-min`, `JS-med`, `JS-min` | Raw diffusion distance/divergence summaries |
| `DSD-min-Rank`, `KL-med-Rank`, `KL-min-Rank`, `JS-med-Rank`, `JS-min-Rank` | Ranks derived from the corresponding raw metrics |

The suffixes describe how target-to-disease relationships are summarized:

- `min` uses the minimum disease-seed distance/divergence for each target before
  averaging across targets;
- `med` uses the median disease-seed value for each target before averaging
  across targets.

The rank columns are calculated in ascending order from the corresponding
non-zero raw distance/divergence values. Therefore:

- **smaller raw DSD/KL/JS values indicate greater network similarity or
  closeness**; and
- **smaller rank values are better**, with rank `1` representing the strongest
  value for that diffusion metric.

Zero raw values are not assigned an ordinary rank in the current backend.

## Main parameters

| Parameter | Interpretation |
|---|---|
| `ties_method` | Method used to resolve ties in diffusion-derived ranks |
| `output_dir` | Directory used to store/reuse diffusion matrices and related intermediate files |

The diffusion matrices can be computationally expensive to generate. Reusing a
stable `output_dir` allows compatible precomputed matrices to be reused in later
runs.

## Main considerations

Diffusion methods capture distributed network similarity beyond immediate
neighbors or shortest paths, but they are more computationally demanding and
depend strongly on the topology of the selected PPI network.

# Method comparison

| Method | Network signal | Main result direction | Main consideration |
|---|---|---|---|
| TrustRank | Disease-seeded random-walk influence | Higher `Score` is better | Sensitive to seed quality and propagation settings |
| Harmonic centrality | Weighted inverse shortest-path accessibility | Higher `Score` is better | Can favor broadly accessible nodes |
| Degree centrality | Penalized 1-hop/2-hop seed-neighborhood connectivity | Higher `Score` is better | Local-topology emphasis |
| Network proximity | Distance to disease module relative to randomized null | More negative `Z` is stronger proximity | Requires informative module/network geometry |
| Diffusion | DSD/KL/JS network similarity | Lower raw metric and lower rank are better | Computationally intensive |

Native values from these methods are not on a common numerical scale. DrugSigNet
therefore converts method-specific outputs to rankings before downstream
consensus analysis rather than directly averaging raw scores.

# Individual method calls

The examples below demonstrate the individual network methods. Full pipeline
orchestration, rank aggregation, annotation, and visualization are covered in
`network-workflow.Rmd`.

## Prepare network inputs

```{r network-example-input}
data("disease_seeds", package = "DrugSigNet")
disease_genes <- disease_seeds

auth_token <- Sys.getenv("SYNAPSE_AUTH_TOKEN")

networks <- load_drugsignet_network(
  network = "all",
  auth_token = auth_token
)

ppi_network <- networks$gene_gene
drug_target_network <- networks$drug_target
```

## TrustRank

```{r run-trustrank}
trust_rank_result <- TrustRank(
  ppi_network = ppi_network,
  drug_target_network = drug_target_network,
  disease_genes = disease_genes,
  hub_penalty = 0.01,
  damping_factor = 0.95
)

head(trust_rank_result@result)
```

## Harmonic centrality

```{r run-harmonic-centrality}
harmonic_result <- Harmonic_centrality(
  ppi_network = ppi_network,
  drug_target_network = drug_target_network,
  disease_genes = disease_genes,
  hub_penalty = 0.01
)

head(harmonic_result@result)
```

## Degree centrality

```{r run-degree-centrality}
degree_result <- Degree_centrality(
  ppi_network = ppi_network,
  drug_target_network = drug_target_network,
  disease_genes = disease_genes
)

head(degree_result@result)
```

## Network proximity

```{r run-network-proximity}
proximity_result <- Network_proximity(
  ppi_network = ppi_network,
  drug_target_network = drug_target_network,
  disease_genes = disease_genes,
  n_simulations = 1000,
  n_workers = 1,
  random_seed = 42
)

head(proximity_result@result)
```

## Diffusion

```{r run-diffusion}
diffusion_result <- Diffusion(
  ppi_network = ppi_network,
  drug_target_network = drug_target_network,
  disease_genes = disease_genes,
  ties_method = "max",
  output_dir = tempdir()
)

head(diffusion_result@result)
```

# Network diagnostics

Network quality can materially affect network-based drug rankings. Before
relying heavily on proximity or diffusion results, inspect whether the disease
and drug-target inputs are suitable for the selected interactome.

Useful checks include:

- the number and fraction of disease seeds mapped to the interactome;
- the number and fraction of drug targets mapped to the interactome;
- the size of the largest connected component;
- the degree distribution of disease seeds;
- disease-seed shortest-path structure and module compactness;
- the number of mapped targets per drug; and
- the extent to which shortest-path distances are compressed in a dense
  interactome.

When network diagnostics indicate weak module structure, poor target coverage,
or limited proximity resolution, centrality-based rankings may be more
informative than relying heavily on proximity or diffusion alone.

# Practical recommendations

- Record the PPI and drug-target networks used for each analysis.
- Retain the `@parameters` slot when archiving a method result because it stores
  the resolved inputs and method-specific settings.
- Check disease-seed and drug-target mapping before interpreting missing or
  weakly ranked candidates.
- Interpret each method according to its own score direction; do not compare raw
  values across different algorithms.
- Compare complementary topology-, distance-, and diffusion-based rankings
  before drawing biological conclusions.
- Use rank aggregation to reduce dependence on any single network statistic.
- Treat highly ranked drugs as computational prioritization hypotheses requiring
  biological interpretation and validation.

# Summary

This vignette described the relevant DrugSigNet concepts, inputs, outputs, and interpretation principles. Preserve the input data, function arguments, package version, and returned S4 objects when reporting or reproducing an analysis.

# Next steps

| Goal | Recommended vignette |
|---|---|
| Choose between signature, network, and integrated analysis | `choosing-analysis-strategy.Rmd` |
| Run the complete network workflow | `network-workflow.Rmd` |
| Understand rank aggregation | `rank-aggregation.Rmd` |
| Annotate prioritized drugs | `drug-annotation.Rmd` |
| Visualize network and consensus results | `visualization.Rmd` |
| Run an integrated example | `complete-case-study.Rmd` |

# Session information

Record the R and package versions used to build this vignette.

```{r session-info, eval=TRUE}
sessionInfo()
```
