---
title: "Network workflow with DrugSigNet"
author: "DrugSigNet authors"
date: "Last update: `r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{09. Network workflow with DrugSigNet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

This vignette shows how to run and inspect the complete network-based DrugSigNet
workflow.

The emphasis is on how method-specific network outputs are converted to ranks,
how diffusion evidence is aggregated, and how topology, proximity, and
diffusion evidence are combined into the final network consensus. For
algorithm-specific details, see `network-methods.Rmd`; for CRank, Dowdall,
and RRA, see `rank-aggregation.Rmd`.

# Workflow overview

`drugNetworkPipeline()` performs the following steps:

1. validates disease genes and resolves the PPI and drug-target networks;
2. runs TrustRank, harmonic centrality, and degree centrality;
3. converts centrality scores to smaller-is-better ranks;
4. optionally runs network proximity and diffusion when
   `run_all_network_methods = TRUE`;
5. converts network-proximity Z-scores to ranks;
6. aggregates the five diffusion rank measures with CRank, Dowdall, and RRA;
7. combines centrality, proximity, and diffusion evidence with CRank, Dowdall,
   and RRA; and
8. returns the final `Network_Harmonized` ranking table for downstream
   annotation and visualization.

The pipeline always runs:

```text
TrustRank
Harmonic_centrality
Degree_centrality
```

With `run_all_network_methods = TRUE`, it additionally runs:

```text
Network_proximity
Diffusion
```

# Prepare disease genes

Disease-associated genes can be supplied as a character vector or a data frame.
When a data frame is used, DrugSigNet expects a `gene` column; otherwise the
first column is converted internally to `gene`.

```{r disease-genes}
data("disease_seeds", package = "DrugSigNet")

disease_genes <- disease_seeds

head(disease_genes)
```

Missing and empty identifiers are removed and duplicated genes are collapsed
before analysis.

Gene identifiers should be compatible with the PPI network and drug-target
network used for the analysis.

# Minimal network run

For an initial analysis, run the three centrality-based methods only.

```{r network-minimal-run}
net_res <- drugNetworkPipeline(
  disease_genes = disease_genes,
  run_all_network_methods = FALSE,
  ties_method = "max",
  prior = 0.093,
  num_bin = 200,
  top_k = 100,
  run_drug_annotation = FALSE,
  run_visualization = FALSE
)
```

When `run_all_network_methods = FALSE`, DrugSigNet runs only:

```text
TrustRank
Harmonic_centrality
Degree_centrality
```

`Network_proximity()` and `Diffusion()` are not run, no diffusion aggregation is
created, and the final network consensus is calculated from the three
centrality-derived ranks only.

# Full network run

Set `run_all_network_methods = TRUE` to add network proximity and diffusion.

```{r network-full-run}
net_res_full <- drugNetworkPipeline(
  disease_genes = disease_genes,
  run_all_network_methods = TRUE,
  hub_penalty = 0.01,
  damping_factor = 0.95,
  n_simulations = 1000,
  n_workers = 5,
  ties_method = "max",
  prior = 0.093,
  num_bin = 200,
  top_k = 100,
  run_drug_annotation = FALSE,
  run_visualization = FALSE
)
```

Centrality methods are run sequentially for Python/reticulate backend stability.
`n_workers` is currently used by network proximity when parallel execution is
supported.

The same workflow can also be called through:

```{r network-unified-run}
net_res_unified <- drugRepurposingPipeline(
  disease_genes = disease_genes,
  mode = "network",
  run_all_network_methods = TRUE,
  top_k = 100,
  run_drug_annotation = FALSE,
  run_visualization = FALSE
)
```

# Network inputs

If `ppi_network` or `drug_target_network` is `NULL`, DrugSigNet resolves the
corresponding default network from its available cache or Synapse-backed
resource.

Custom networks can be supplied directly.

The PPI network should contain:

```text
gene1
gene2
```

The drug-target network should contain:

```text
ID
Drug
Target
Group
```

For example:

```{r custom-networks}
ppi_network <- data.frame(
  gene1 = c("APOE", "APOE", "CLU", "APP"),
  gene2 = c("CLU", "APP", "BIN1", "PSEN1"),
  stringsAsFactors = FALSE
)

drug_target_network <- data.frame(
  ID = c("D1", "D2", "D3"),
  Drug = c("drug_a", "drug_b", "drug_c"),
  Target = c("APOE", "APP", "PSEN1"),
  Group = c("approved", "approved", "approved"),
  stringsAsFactors = FALSE
)

net_custom <- drugNetworkPipeline(
  ppi_network = ppi_network,
  drug_target_network = drug_target_network,
  disease_genes = disease_genes,
  run_all_network_methods = FALSE,
  run_drug_annotation = FALSE,
  run_visualization = FALSE
)
```

The PPI endpoints, disease genes, and drug targets should use compatible gene
identifiers.

# Important arguments

| Argument | Meaning |
|---|---|
| `disease_genes` | Disease-associated seed genes |
| `ppi_network` | Optional custom PPI network with `gene1` and `gene2` |
| `drug_target_network` | Optional custom drug-target table |
| `run_all_network_methods` | Add network proximity and diffusion when `TRUE` |
| `include_indirect_drugs` | Include indirectly connected drugs |
| `include_non_approved_drugs` | Include non-approved or investigational drugs |
| `hub_penalty` | Hub penalty for TrustRank and harmonic centrality |
| `damping_factor` | TrustRank propagation parameter; default `0.95` |
| `result_size` | Number of ranked drugs requested from network methods; when `NULL`, inferred from the resolved drug-target network |
| `n_simulations` | Random simulations used by network proximity |
| `n_workers` | Workers used by supported parallel network methods |
| `prior` | CRank prior; default `0.093` |
| `num_bin` | Number of CRank rank intervals; default `200` |
| `ties_method` | Tie-handling rule used during rank conversion and aggregation |
| `output_dir` | Optional directory for diffusion intermediate files |
| `top_k` | Top-rank threshold used for downstream annotation/visualization |


# Inspect the result structure

`drugNetworkPipeline()` returns a `DrugSearchingPipeline` object.

```{r inspect-network-structure}
slotNames(net_res_full)

names(net_res_full@DrugSearching)
names(net_res_full@RankAggregation)
```

The principal network sections are:

| Section | Contents |
|---|---|
| `@DrugSearching[["Raw"]]` | Native `NetworkBased` outputs |
| `@DrugSearching[["Processed"]]` | Method outputs with pipeline-ready rank information |
| `@RankAggregation[["Diffusion"]]` | CRank, Dowdall, and RRA consensus across diffusion rank measures; present only when `run_all_network_methods = TRUE` |
| `@RankAggregation[["Harmonized"]]` | Final CRank, Dowdall, and RRA network aggregation objects |
| `@RankAggregation[["Network_Harmonized"]]` | Final analysis-ready network ranking table |

When `run_all_network_methods = FALSE`, the raw and processed sections contain
only TrustRank, harmonic centrality, and degree centrality, and
`@RankAggregation[["Diffusion"]]` is absent/`NULL`.

# Raw network-method results

Inspect the available raw methods with:

```{r inspect-network-raw}
names(net_res_full@DrugSearching[["Raw"]])

head(
  net_res_full@DrugSearching[["Raw"]][["TrustRank"]]@result
)

head(
  net_res_full@DrugSearching[["Raw"]][["Network_proximity"]]@result
)

head(
  net_res_full@DrugSearching[["Raw"]][["Diffusion"]]@result
)
```

Each raw entry is a `NetworkBased` object with `@result` and method-specific
`@parameters`.

The important native score directions are:

| Method | Native evidence | Better direction |
|---|---|---|
| TrustRank | `Score` | Higher |
| Harmonic centrality | `Score` | Higher |
| Degree centrality | `Score` | Higher |
| Network proximity | `Z` | More negative / smaller |
| Diffusion | DSD, KL, JS metrics and their ranks | Lower metric / lower rank |

Detailed method definitions and raw result columns are documented in
`network-methods.Rmd`.

# Processed network rankings

The pipeline creates a processed representation for each `NetworkBased` result.

```{r inspect-network-processed}
names(net_res_full@DrugSearching[["Processed"]])

head(
  net_res_full@DrugSearching[["Processed"]][["TrustRank"]]
)

head(
  net_res_full@DrugSearching[["Processed"]][["Network_proximity"]]
)
```

For centrality methods, the native larger-is-better `Score` is converted to a
smaller-is-better `Rank`:

```text
highest Score -> Rank 1
next Score    -> Rank 2
...
```

For network proximity, smaller Z-scores are ranked first, so the most negative
Z-score receives the strongest rank.

Diffusion already contains method-specific rank columns and therefore does not
require a single additional `Rank` column.

# Centrality ranks used for aggregation

The three centrality result tables are joined by drug, and their `Score`
columns are converted to ranks before consensus aggregation.

The resulting aggregation inputs are conceptually:

```text
Drug
TrustRank
Harmonic_centrality
Degree_centrality
```

At this stage these columns contain **ranks**, not the original centrality
scores, and smaller values indicate stronger priority.

This distinction is important when interpreting the final
`Network_Harmonized` table, where the centrality columns retain their method
names but contain rank values.

# Network-proximity rank

When the full network workflow is enabled, the proximity result is converted
from its native Z-score to a single rank column:

```text
Network_proximity
```

The rank is calculated with smaller Z-scores first. Therefore:

- more negative `Z` = stronger native proximity evidence;
- smaller `Network_proximity` rank = stronger priority.

The raw `D`, `Z`, `Mean`, and `SD` values remain available in
`@DrugSearching[["Raw"]][["Network_proximity"]]@result`.

# Diffusion rank aggregation

The diffusion output contributes five rank measures:

```text
DSD-min-Rank
KL-med-Rank
KL-min-Rank
JS-med-Rank
JS-min-Rank
```

Lower values indicate stronger diffusion-based similarity.

Before diffusion is combined with the other network methods, these five rank
columns are aggregated independently with CRank, Dowdall, and RRA.

```{r inspect-diffusion-aggregation}
names(
  net_res_full@RankAggregation[["Diffusion"]]
)

net_res_full@RankAggregation[["Diffusion"]][["CRank"]]@result
net_res_full@RankAggregation[["Diffusion"]][["Dowdall"]]@result
net_res_full@RankAggregation[["Diffusion"]][["RRA"]]@result
```

The resulting columns are:

```text
Diffusion_CRank
Diffusion_Dowdall
Diffusion_RRA
```

Only complete rows across the available diffusion rank columns are passed to
this diffusion-aggregation stage.

Because the diffusion inputs are already smaller-is-better ranks, the pipeline
uses:

```text
CRank:   reverse = TRUE
Dowdall: reverse = FALSE
RRA:     reverse = FALSE
```

# Harmonizing network evidence

The final aggregation stage combines the available network evidence.

## Centrality-only workflow

With:

```r
run_all_network_methods = FALSE
```

the pipeline stops after the three centrality methods have been converted to
ranks. No network-proximity or diffusion result is added.

Each final aggregation method therefore combines only:

```text
TrustRank
Harmonic_centrality
Degree_centrality
```

These columns are smaller-is-better ranks. The resulting consensus columns are
still:

```text
Network_CRank
Network_Dowdall
Network_RRA
```

## Full network workflow

With:

```r
run_all_network_methods = TRUE
```

the three centrality ranks are combined with:

```text
Network_proximity
```

and the diffusion consensus corresponding to the same aggregation method.

Therefore the three final aggregation inputs are conceptually:

| Final consensus | Inputs |
|---|---|
| `Network_CRank` | Three centrality ranks + `Network_proximity` + `Diffusion_CRank` |
| `Network_Dowdall` | Three centrality ranks + `Network_proximity` + `Diffusion_Dowdall` |
| `Network_RRA` | Three centrality ranks + `Network_proximity` + `Diffusion_RRA` |

This preserves the aggregation framework used to summarize the diffusion
metrics before combining diffusion with the other network evidence.

All harmonized inputs are smaller-is-better ranks, so the pipeline again uses:

```text
CRank:   reverse = TRUE
Dowdall: reverse = FALSE
RRA:     reverse = FALSE
```

Inspect the aggregation objects with:

```{r inspect-network-harmonized}
names(
  net_res_full@RankAggregation[["Harmonized"]]
)

net_res_full@RankAggregation[["Harmonized"]][["CRank"]]@result
net_res_full@RankAggregation[["Harmonized"]][["Dowdall"]]@result
net_res_full@RankAggregation[["Harmonized"]][["RRA"]]@result
```

# Final network ranking table

The final analysis-ready table is:

```{r inspect-network-harmonized-table}
network_rankings <-
  net_res_full@RankAggregation[["Network_Harmonized"]]

head(network_rankings)
colnames(network_rankings)
```

When `run_all_network_methods = FALSE`, `Network_Harmonized` contains the
three centrality ranks together with:

```text
Network_CRank
Network_Dowdall
Network_RRA
```

These final consensus ranks are calculated only from TrustRank, harmonic
centrality, and degree centrality.

When `run_all_network_methods = TRUE`, the table additionally retains the
network-proximity rank, the five diffusion rank columns, and the three
diffusion consensus ranks. In that case, proximity and the corresponding
diffusion consensus are also included in each final network aggregation.

The final network consensus columns are ranks, so **smaller values indicate
higher network-based priority**.

For example:

```{r top-network-consensus}
network_rankings |>
  dplyr::arrange(Network_CRank) |>
  dplyr::select(
    Drug,
    Network_CRank,
    Network_Dowdall,
    Network_RRA
  ) |>
  head(20)
```

Agreement across the three consensus formulations can be used to assess the
robustness of candidate prioritization without assuming that one aggregation
method is universally preferable.

# Downstream annotation and visualization

Annotation and visualization can be enabled directly in the network pipeline
with:

```text
run_drug_annotation = TRUE
run_visualization = TRUE
```

For downstream annotation, DrugSigNet selects the union of drugs ranked within
the top `top_k` by `Network_CRank`, `Network_Dowdall`, or `Network_RRA`.

When annotation is enabled and drug targets are available, targets of those
top-ranked drugs are also used for GO target-set enrichment with `TSEA()`.

See `drug-annotation.Rmd` and `visualization.Rmd` for the standalone
annotation and plotting interfaces.

# Exporting and reporting results

Completed network pipeline results can be exported with
`write_pipeline_results()` or summarized in an HTML/PDF report with
`write_report()`. See `complete-case-study.Rmd` for complete examples.

# Interpretation checklist

- What fraction of disease genes map to the PPI network?
- Are drug targets sufficiently represented in the network?
- Are the centrality rankings broadly consistent or complementary?
- Does network proximity support the same drugs as topology-based methods?
- Are diffusion results stable and computationally feasible for the network?
- Do `Network_CRank`, `Network_Dowdall`, and `Network_RRA` identify a stable
  high-priority candidate set?
- Are the prioritized drugs supported by biologically plausible targets?

# Troubleshooting

## Few disease genes map to the network

Check the identifier system used by the disease genes and both network inputs.
The PPI endpoints, disease seeds, and drug targets must be mutually compatible.

## Network proximity has limited coverage

Inspect drug-target mapping and the number of mapped targets per drug. Small
target sets or incomplete network coverage can reduce the information available
to randomization-based proximity analysis.

## Diffusion requires substantial memory

Diffusion can require substantial memory because the current implementation
uses large network-derived matrices and intermediate files. If computational
resources are limited, validate the centrality-only workflow first and enable
diffusion only for the full analysis.

## Full workflow is slow

Start with `run_all_network_methods = FALSE` to validate the disease genes,
network mapping, and centrality results. Add proximity and diffusion only when
those inputs and computational requirements are appropriate.

# Summary

This vignette described the relevant DrugSigNet concepts, inputs, outputs, and interpretation principles. Preserve the input data, function arguments, package version, and returned S4 objects when reporting or reproducing an analysis.

# Next steps

| Goal | Recommended vignette |
|---|---|
| Understand individual network methods | `network-methods.Rmd` |
| Understand CRank, Dowdall, and RRA | `rank-aggregation.Rmd` |
| Annotate prioritized drugs | `drug-annotation.Rmd` |
| Visualize network rankings | `visualization.Rmd` |
| Compare with the signature workflow | `signature-workflow.Rmd` |
| Integrate signature and network evidence | `complete-case-study.Rmd` |

# Session information

Record the R and package versions used to build this vignette.

```{r session-info, eval=TRUE}
sessionInfo()
```
