---
title: "Rank aggregation with DrugSigNet"
author: "DrugSigNet authors"
date: "Last update: `r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{05. Rank aggregation with DrugSigNet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

DrugSigNet uses rank aggregation to combine evidence from complementary
signature-based, network-based, and integrated drug-prioritization methods.

This vignette explains the three aggregation methods implemented in DrugSigNet:
CRank, Dowdall aggregation, and Robust Rank Aggregation (RRA). The focus is on
input direction, method-specific outputs, interpretation, parameter provenance,
and reproducible standalone use. For complete signature, network, and integrated
workflows, see the corresponding workflow vignettes.

# Why aggregate rankings?

Different drug-prioritization methods can disagree because they encode different
biological and statistical assumptions. Signature methods may emphasize
gene-set enrichment or quantitative expression similarity, while network
methods may emphasize local topology, shortest-path distance, propagation, or
diffusion.

Rank aggregation combines these complementary signals into consensus
prioritizations and reduces dependence on any single method.

Consensus should not be interpreted as experimental validation. A high
consensus rank indicates convergence of computational evidence and should be
followed by biological interpretation and, where appropriate, experimental or
clinical validation.

# Input requirements

All three DrugSigNet aggregation functions expect a data frame in which the
first column identifies the item, typically a drug, and at least two additional
columns are numeric ranking variables.

For example:

```{r rank-aggregation-worked-input}
rank_input <- data.frame(
  Drug = paste0("drug_", letters[1:10]),
  Method1 = c(1, 3, 2, 7, 5, 4, 10, 6, 8, 9),
  Method2 = c(2, 1, 4, 3, 6, 8, 5, 10, 7, 9),
  Method3 = c(3, 2, 1, 5, 4, 7, 6, 8, 10, 9),
  stringsAsFactors = FALSE
)
```

In this example, the numeric columns already contain ranks, so smaller values
are better.

# Input direction and `reverse`

Correct direction handling is essential because the three aggregation functions
do not use the same `reverse` convention.

| Meaning of the numeric input columns | `CRank()` | `Dowdall()` | `RRA()` |
|---|---:|---:|---:|
| Smaller values are better, for example ranks `1, 2, 3, ...` | `reverse = TRUE` | `reverse = FALSE` | `reverse = FALSE` |
| Larger values are better, for example DrugSigNet `rank_score` values | `reverse = FALSE` | `reverse = TRUE` | `reverse = TRUE` |

Therefore, do not copy the same `reverse` value across the three methods without
first checking what the input columns mean.

For the rank-based example above, the correct calls are:

```{r direction-example}
crank_result <- CRank(
  input_data = rank_input,
  reverse = TRUE
)

dowdall_result <- Dowdall(
  input_data = rank_input,
  reverse = FALSE
)

rra_result <- RRA(
  input_data = rank_input,
  reverse = FALSE
)
```

By contrast, when aggregating larger-is-better DrugSigNet `rank_score` columns,
the corresponding settings are:

```text
CRank:   reverse = FALSE
Dowdall: reverse = TRUE
RRA:     reverse = TRUE
```

# Inspecting results and parameters

Each aggregation method returns a `RankAggregation` S4 object containing both the
method-specific consensus result and the input/settings used to generate it.

| Slot | Contents |
|---|---|
| `@result` | Method-specific consensus results |
| `@parameters` | Original aggregation input and retained method settings |

The parameter slot is useful for reproducibility because it preserves the exact
input table supplied to the aggregation function.

For example, an aggregation stored inside a DrugSigNet pipeline can be inspected
with:

```{r inspect-rank-aggregation-parameters}
agg <- covid_rep@RankAggregation[["signature"]][["Pairwise"]][["CMAP"]][["CRank"]]

names(agg@parameters)
agg@parameters[["input_data"]]
```

For CRank, retained provenance can include `input_data`, `rank_aggregation`,
`ties_method`, `prior`, `num_bin`, `num_iter`, `reverse`, `converged`, and
`correlation`. The exact fields differ by aggregation method. In particular,
CRank additionally records convergence information from the iterative
procedure.

# CRank

## How the method works

CRank is an iterative Bayesian consensus-ranking framework. The numeric input
columns are first converted to ranks according to the selected direction and
tie-handling rule.

CRank then normalizes the ranks, partitions them into rank intervals, and
iteratively estimates which rank regions provide evidence for membership in a
temporary high-confidence set. Rank intervals that agree more strongly with the
current high-confidence set receive greater influence in the next consensus
iteration.

The procedure continues until convergence or until the maximum number of
iterations is reached.

## Main parameters

| Parameter | Meaning |
|---|---|
| `prior` | Fraction of items used to define the temporary high-confidence group |
| `num_bin` | Number of rank intervals used for Bayesian evidence estimation |
| `num_iter` | Maximum number of CRank iterations |
| `ties_method` | Method used to resolve ties during ranking |
| `reverse` | Defines whether smaller or larger input values represent stronger evidence |

DrugSigNet commonly uses:

```text
prior = 0.093
num_bin = 200
num_iter = 1000
ties_method = "max"
```

The temporary high-confidence group size is determined internally from the
number of input items and `prior`.

## Result columns

A CRank result contains:

```text
Drug    CRank
```

| Column | Meaning |
|---|---|
| `Drug` | Drug or item identifier |
| `CRank` | Final CRank consensus rank |

**Smaller `CRank` values indicate higher consensus priority.**

For example, an output can look like:

```text
Drug              CRank
sulindac             764
fludrocortisone     1800
latamoxef             162
colchicine            727
amitriptyline         336
mesalazine            980
```

The returned rows are not guaranteed to be sorted by the final `CRank` value.
Therefore, `head()` should not automatically be interpreted as the top-ranked
CRank candidates.

To inspect the highest-priority drugs, explicitly sort the result:

```{r sort-crank}
crank_result@result |>
  dplyr::arrange(CRank) |>
  head()
```

## Convergence

CRank stores convergence information in `@parameters`. It is useful to inspect
these fields when documenting or auditing an analysis:

```{r inspect-crank-convergence}
crank_result@parameters[["converged"]]
crank_result@parameters[["correlation"]]
```

For publication-oriented analyses, retain the CRank parameter object together
with the result so that the input, parameterization, and convergence status can
be recovered.

## Missing values and zero entries

CRank replaces missing numeric values with zero before its internal rank
conversion. Zero values are treated as unranked or zero-evidence entries rather
than as ordinary positive ranking values.

This convention is particularly relevant when DrugSigNet constructs a common
drug universe across multiple methods and some drugs are unavailable or
unsupported in one of the input lists.

# Dowdall aggregation

## How the method works

DrugSigNet's Dowdall implementation is based on reciprocal-rank voting.

For each numeric input column, the method:

1. converts non-zero input values to ranks using the selected direction;
2. converts each non-zero rank to `1 / rank`;
3. min-max scales the reciprocal-rank contributions within that column to the
   range `[0, 1]`; and
4. sums the scaled contributions across methods.

Drugs receiving strong support near the top of several lists therefore obtain
larger consensus scores.

## Result columns

A typical DrugSigNet Dowdall result contains the item identifier, transformed
input columns, `Dowdall_score`, and `Dowdall_rank`.

For example:

```text
Drug             rank_score_1  rank_score_2  Dowdall_score  Dowdall_rank
sulindac               1          0.00108          1.00             1
fludrocortisone        0.5        0                0.5              4
latamoxef              0.333      0.000618         0.334            6
colchicine             0.25       0.00104          0.251            8
amitriptyline          0.2        0.000756         0.201            9
mesalazine             0.167      0.00138          0.168           11
```

The result columns should be interpreted as follows:

| Column | Meaning |
|---|---|
| `Drug` | Drug or item identifier |
| numeric input columns | Scaled reciprocal-rank contributions after Dowdall transformation |
| `Dowdall_score` | Sum of the transformed contributions across methods |
| `Dowdall_rank` | Final rank derived from `Dowdall_score` |

For interpretation, **higher `Dowdall_score` is better**, whereas **smaller
`Dowdall_rank` is better**.

As with CRank, explicitly sorting by the final rank is the safest way to inspect
the highest-priority drugs:

```{r sort-dowdall}
dowdall_result@result |>
  dplyr::arrange(Dowdall_rank) |>
  head()
```

## Missing values and zero entries

Missing numeric values are replaced with zero. Zero entries are treated as
unranked and contribute zero to the final consensus score.

# Robust Rank Aggregation

## How the method works

Robust Rank Aggregation (RRA) evaluates whether an item appears consistently
near the top of multiple ranked lists more often than expected under a null
model of random ordering.

DrugSigNet first converts the numeric input columns to ranks using the requested
direction and tie-handling rule, normalizes the rank matrix, and then calculates
a rho score for each item using `RobustRankAggreg::rhoScores()`.

The resulting score is reported as `RRA_pval`.

## Result columns

A typical RRA result contains:

```text
Drug          RRA_pval    RRA_rank
letrozole     0.001669136        1
sulindac      0.002221605        3
oxypurinol    0.002221605        3
mephenytoin   0.002688889        4
ozagrel       0.003200000        5
chlorfenson   0.003472222        6
```

| Column | Meaning |
|---|---|
| `Drug` | Drug or item identifier |
| `RRA_pval` | RRA rho/significance score |
| `RRA_rank` | Rank derived from `RRA_pval` |

For interpretation, **smaller `RRA_pval` is better** and **smaller `RRA_rank`
is better**.

Ties in `RRA_pval` can therefore produce tied final ranks, depending on
`ties_method`.

## Missing values

Unlike CRank and Dowdall, RRA preserves missing values during rank conversion.
Missing values are not converted to zero before RRA scoring.

This distinction is important when the input ranking lists are incomplete.

# Method comparison

| Method | Consensus principle | Main output | Better direction | Missing-value handling |
|---|---|---|---|---|
| CRank | Iterative Bayesian weighting of rank regions | `CRank` | Lower | Numeric `NA` replaced by `0`; zero treated as unranked |
| Dowdall | Scaled reciprocal-rank voting | `Dowdall_score`, `Dowdall_rank` | Higher score; lower rank | Numeric `NA` replaced by `0`; zero contributes no evidence |
| RRA | Unexpected top-rank consistency under a null model | `RRA_pval`, `RRA_rank` | Lower p-value; lower rank | Missing values preserved during rank conversion |

The native outputs are not numerically interchangeable. For example, a CRank
value, a Dowdall score, and an RRA p-value have different meanings even when
they refer to the same drug.

DrugSigNet therefore retains method-specific aggregation outputs rather than
treating the three values as measurements on a common scale.

# How DrugSigNet uses rank aggregation

DrugSigNet applies CRank, Dowdall, and RRA at several levels of the workflow.

## Signature-based aggregation

Within a signature-search family, rankings generated against different
reference databases can first be aggregated. DrugSigNet can then harmonize
evidence across CMAP, LINCS, gCMAP, and correlation families.

For processed signature `rank_score` inputs, larger values represent stronger
evidence. The aggregation direction is therefore handled accordingly:

```text
CRank:   reverse = FALSE
Dowdall: reverse = TRUE
RRA:     reverse = TRUE
```

## Network-based aggregation

Centrality outputs are first converted to ranks. When the full network workflow
is used, network-proximity evidence and diffusion-based ranks are also
incorporated.

The five diffusion rank columns can themselves be aggregated using CRank,
Dowdall, and RRA before they are incorporated into the broader network
consensus.

Because centrality scores are larger-is-better while proximity, diffusion, and
consensus tables are typically represented as smaller-is-better ranks, DrugSigNet
handles direction before applying each aggregation method.

## Integrated aggregation

The integrated DrugSigNet workflow does not simply average the final signature
and network consensus columns.

Instead, method-specific evidence from the signature and network analyses is
assembled into integration inputs and CRank, Dowdall, and RRA are applied again
to generate integrated consensus rankings.

# Run the rank aggregation methods

```{r rank-aggregation-input}
rank_input <- data.frame(
  Drug = paste0("drug_", letters[1:10]),
  CMAP = c(1, 3, 2, 7, 5, 4, 10, 6, 8, 9),
  LINCS = c(2, 1, 4, 3, 6, 8, 5, 10, 7, 9),
  Correlation = c(3, 2, 1, 5, 4, 7, 6, 8, 10, 9),
  stringsAsFactors = FALSE
)
```

The first column identifies the items being aggregated. Each remaining numeric
column is one method-specific ranking, where smaller values are better. Therefore
all three calls use `reverse = FALSE`.

## Dowdall

```{r run-dowdall}
dowdall_result <- Dowdall(
  input_data = rank_input,
  ties_method = "max",
  reverse = FALSE
)
dowdall_result@result
```

## Robust Rank Aggregation

```{r run-rra}
rra_result <- RRA(
  input_data = rank_input,
  full = TRUE,
  exact = FALSE,
  ties_method = "max",
  reverse = FALSE
)
rra_result@result
```

## CRank

CRank uses the package's Python-backed implementation. Configure the Python
dependencies before running this section.

```{r run-crank}
crank_result <- CRank(
  input_data = rank_input,
  ties_method = "max",
  prior = 0.093,
  num_bin = 200,
  num_iter = 1000,
  reverse = FALSE
)
crank_result@result
```
# Reusing stored pipeline inputs

Because the original aggregation input is retained in `@parameters`, an existing
pipeline aggregation can be rerun directly.

For example:

```{r rerun-stored-input}
stored_input <-
  covid_rep@RankAggregation[["signature"]][["Pairwise"]][["CMAP"]][["CRank"]]@
    parameters[["input_data"]]

crank_check <- CRank(
  input_data = stored_input,
  reverse = FALSE
)
```

The appropriate `reverse` value must be chosen according to the semantics of the
stored numeric columns. In this example, DrugSigNet pairwise signature
`rank_score` columns are larger-is-better, so CRank uses `reverse = FALSE`.

The equivalent larger-is-better input would use:

```{r rerun-stored-input-other-methods}
dowdall_check <- Dowdall(
  input_data = stored_input,
  reverse = TRUE
)

rra_check <- RRA(
  input_data = stored_input,
  reverse = TRUE
)
```

# Tie handling

DrugSigNet supports several tie-handling rules through `ties_method`. Common
choices include `"max"`, `"min"`, `"average"`, and `"dense"`.

The choice affects both the intermediate rank conversion and, depending on the
aggregation method, the final consensus rank.

For reproducible analyses, retain the same tie-handling rule across comparable
runs unless there is a specific reason to change it.

When a top-k set is defined from ranks, ties can result in more than exactly
`k` drugs satisfying a criterion such as `rank <= k`.

# Practical reporting

For a reproducible DrugSigNet rank-aggregation analysis, report:

- the input ranking columns and whether larger or smaller values represented
  stronger evidence;
- the aggregation methods used;
- `reverse` and `ties_method` for each aggregation call;
- CRank `prior`, `num_bin`, and `num_iter`;
- CRank convergence status where relevant;
- the treatment of missing or unranked entries;
- the number of items included in the aggregation;
- the top-k rule used for downstream prioritization; and
- agreement or overlap across CRank, Dowdall, and RRA outputs.

# Common pitfalls

## Using the wrong `reverse` setting

The `reverse` convention differs between CRank and Dowdall/RRA. Always determine
whether the input is smaller-is-better or larger-is-better before calling the
method.

## Interpreting `head()` as the top-ranked drugs

CRank and Dowdall outputs are not guaranteed to be returned in final-rank order.
Explicitly arrange by `CRank`, `Dowdall_rank`, or `RRA_rank` before interpreting
the leading candidates.

## Mixing scores and ranks

A larger-is-better score and a smaller-is-better rank require different
direction handling. Do not aggregate them without first confirming their
semantics.

## Ignoring missing-value conventions

CRank and Dowdall convert numeric missing values to zero, whereas RRA preserves
missingness during rank conversion. This can matter when input lists have
different coverage.

## Aggregating incompatible drug universes

Harmonize drug names and identifiers before aggregation. Differences caused only
by inconsistent naming can create artificial disagreement or missingness.

## Treating consensus as validation

Consensus ranking summarizes agreement among computational methods; it does not
establish therapeutic efficacy.

# Summary

This vignette described the relevant DrugSigNet concepts, inputs, outputs, and interpretation principles. Preserve the input data, function arguments, package version, and returned S4 objects when reporting or reproducing an analysis.

# Next steps

| Goal | Recommended vignette |
|---|---|
| Choose an overall DrugSigNet strategy | `choosing-analysis-strategy.Rmd` |
| Understand signature-search inputs to aggregation | `signature-methods.Rmd` |
| Understand network-method inputs to aggregation | `network-methods.Rmd` |
| Run the complete signature workflow | `signature-workflow.Rmd` |
| Run the complete network workflow | `network-workflow.Rmd` |
| Run integrated signature-network prioritization | `complete-case-study.Rmd` |

# Session information

Record the R and package versions used to build this vignette.

```{r session-info, eval=TRUE}
sessionInfo()
```
