---
title: "Signature-based methods in DrugSigNet"
author: "DrugSigNet authors"
date: "Last update: `r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{03. Signature-based methods in DrugSigNet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

This vignette explains the signature-based search methods used by DrugSigNet:
CMAP, LINCS, gCMAP, and correlation-based search. It focuses on the biological
and statistical interpretation of these methods, the expression representations
they require, the CMAP and LINCS2 reference databases used by DrugSigNet, and
the native scores returned by each search.

DrugSigNet builds on gene-expression signature search functionality implemented
in the Bioconductor
[`signatureSearch`](https://bioconductor.org/packages/signatureSearch/) package.
For a practical end-to-end DrugSigNet analysis, see `signature-workflow.Rmd`.

# Biological rationale

Signature-based drug repurposing represents both a disease state and a drug
perturbation as molecular expression signatures. A disease signature describes
genes whose expression changes in association with the disease or phenotype,
whereas a drug signature describes expression changes induced by a compound in
a cellular system.

A signature search asks whether a drug-induced profile is related to the query
disease profile. The relationship can be:

- **positive or concordant**, when the drug signature resembles the disease
  signature; or
- **negative or discordant**, when the drug signature opposes the disease
  signature.

Reversal is a common therapeutic hypothesis because a drug that opposes
disease-associated expression changes may counteract part of the disease state.
However, reversal should not be assumed to be therapeutically preferable in
every biological setting. DrugSigNet therefore retains connectivity direction
for interpretation, and the appropriate direction should be considered in the
context of the disease state and analysis objective.

# DrugSigNet signature input

The main DrugSigNet signature workflow uses a data frame containing `Entrez` and
`FC` columns:

- `Entrez` contains Entrez gene identifiers;
- `FC` contains signed fold-change values, commonly log2 fold changes.

The four search methods use two internal representations derived from this
signed profile.

| DrugSigNet method | Query representation used for the search | Reference representation used by the method |
|---|---|---|
| CMAP | Up- and down-regulated gene sets | Ranked reference expression profile |
| LINCS | Up- and down-regulated gene sets | Quantitative differential-expression scores |
| gCMAP | Ranked or quantitative one-column query profile | Signed reference gene sets generated from upper/lower score thresholds |
| Correlation | Quantitative one-column query profile | Quantitative reference expression profile |

For the individual method wrappers, CMAP and LINCS use `upset` and `downset`
gene sets. gCMAP and correlation use a one-column numeric matrix with Entrez
identifiers as row names. `drugSignaturePipeline()` prepares the required
representations internally from `signature_input`.

The upstream `signatureSearch` package also implements Fisher's exact-test
search. Fisher search is not one of the four core signature-search methods
orchestrated by the DrugSigNet signature pipeline.

# Reference databases

## CMAP and LINCS2 resources

DrugSigNet searches drug-induced expression signatures stored in HDF5 reference
databases. The CMAP and LINCS2 reference files correspond to databases provided
by the Bioconductor
[`signatureSearchData`](https://bioconductor.org/packages/signatureSearchData/)
package through `ExperimentHub`.

The frozen HDF5 copies used by the DrugSigNet reference-data workflow preserve
these underlying reference resources so that analyses can be repeated against
the same database files.

| Reference database | Expression statistic | Genes | Compound treatments | Cell lines | Signatures | ExperimentHub ID |
|---|---|---:|---:|---:|---:|---|
| CMAP | Log2 fold-change values | 12,403 | 1,281 | 5 | 3,478 | `EH3223` |
| LINCS2 | Moderated z-scores from differential-expression analysis | 12,328 | 30,456 | 58 | 136,460 | `EH7297` |

The table follows the `signatureSearchData` data-package documentation for the
CMAP HDF5 resource (`EH3223`), which reports 12,403 genes. Some versions of the
upstream `signatureSearch` workflow vignette list 12,437 genes in a summary of
the same resource; recording the ExperimentHub identifier together with the
database statistics therefore provides the clearest provenance.

The CMAP HDF5 database is derived from Connectivity Map/CMAP2 perturbation
profiles. The database contains drug-induced log2 fold-change signatures and is
much smaller than LINCS2 in both compound and cellular coverage.

LINCS2 is derived from the expanded LINCS L1000 collection. The L1000 assay
directly measures a set of landmark genes and infers additional genes. In the
LINCS2 reference used by `signatureSearchData`, an exemplar signature is selected
for each perturbagen within each cell line to reduce redundancy across treatment
conditions. The exemplar selection is based on CLUE annotations and generally
favours signatures with high transcriptional activity scores.

The two databases therefore differ substantially in scale, assay technology,
cellular coverage, and expression statistic. Differences between results
obtained from CMAP and LINCS2 are expected and should not be interpreted as
technical inconsistency by themselves.

For the upstream construction details, see the
[`signatureSearch` vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/signatureSearch/inst/doc/signatureSearch.html)
and the `signatureSearchData` vignette.

## ExperimentHub provenance

The original HDF5 resources can be resolved through `ExperimentHub` as follows.
This is useful for provenance checking; normal DrugSigNet analyses do not need
to retrieve them manually when a supported `ref_db` value is used.

```{r experimenthub-provenance}
library(ExperimentHub)

eh <- ExperimentHub()

cmap_path <- eh[["EH3223"]]
lincs2_path <- eh[["EH7297"]]
```

## Selecting `ref_db`

The individual DrugSigNet signature methods accept a `ref_db` argument.

Use:

- `"cmap"` for the CMAP HDF5 reference;
- `"lincs2"` for the LINCS2 HDF5 reference; or
- a local HDF5 path returned by `load_signature_refdb()` when using a frozen
  DrugSigNet reference database.

A local frozen reference is useful when an analysis needs to be repeated against
an unchanged database file rather than resolving the reference dynamically.

The HDF5 databases are searched in chunks rather than being loaded completely
into memory. The `chunk_size` argument controls the number of reference
signatures processed per chunk and is therefore mainly a computational
performance and memory-management parameter.

# Methods and reference databases are separate choices

The search algorithm and the reference database represent two different parts of
the analysis. DrugSigNet can apply each of its four signature-search methods to
either CMAP or LINCS2, producing eight method/reference combinations in a full
multi-method search:

| Reference | CMAP method | LINCS method | gCMAP method | Correlation method |
|---|---|---|---|---|
| CMAP | `CMAP_CMAP` | `LINCS_CMAP` | `gCMAP_CMAP` | `Correlation_CMAP` |
| LINCS2 | `CMAP_LINCS2` | `LINCS_LINCS2` | `gCMAP_LINCS2` | `Correlation_LINCS2` |

This distinction is important when interpreting disagreement between rankings.
Two results may differ because the **search algorithm** differs, because the
**reference database** differs, or because both differ.

# Inspecting method results and parameters

Each individual DrugSigNet signature search stores the search output in
`@result` and the corresponding query and settings in `@parameters`.

For example, a CMAP search stored in the raw pipeline results can be inspected
with:

```{r inspect-signature-parameters}
names(
  covid_signature@DrugSearching[["Raw"]][["CMAP_CMAP"]]@parameters
)
#> [1] "query" "signature_method" "refdb" "chunk_size"
```

For this search, `@parameters` records the query used, the signature-search
method, the reference database, and the chunk size. This information is useful
for documenting and reproducing the exact method/reference search.

The contents of `@parameters` can vary across signature methods because their
query representations and method-specific settings differ. In general, users
can inspect the parameter provenance of any stored raw search with:

```{r inspect-generic-signature-parameters}
result@DrugSearching[["Raw"]][["METHOD_REFERENCE"]]@parameters
```

The individual method examples below therefore focus on `@result`; the
parameter slot only needs to be inspected when the settings or provenance of a
particular search are required.

# CMAP search

## How the method works

The CMAP search uses two query gene sets representing the most up- and
down-regulated genes. These sets are compared with ranked drug-induced
expression profiles in the reference database.

The method evaluates whether query genes occur preferentially toward the upper
or lower ends of each ranked reference profile and combines the two directions
into a bidirectional connectivity score. The `signatureSearch` implementation
closely follows the original Connectivity Map search formulation.

A positive score indicates concordant connectivity between the query and
reference signature, whereas a negative score indicates discordant
connectivity.

## Native scores

The main CMAP result fields are:

- `raw_score`: bidirectional enrichment/connectivity score;
- `scaled_score`: `raw_score` scaled separately within its positive and negative
  ranges to values between -1 and 1;
- `trend`: direction of connectivity (`up` for positive and `down` for negative
  connectivity in the upstream result representation);
- `N_upset` and `N_downset`: numbers of query genes retained in the corresponding
  up and down sets.

## Main considerations

CMAP is intuitive and relatively technology-agnostic because the query is
reduced to gene sets. This also means that the exact magnitude of the query fold
changes is not directly retained after the up/down sets have been defined.
Results can therefore depend on how the query genes are selected and on how many
genes map to the reference database.

# LINCS search

## How the method works

The LINCS method also uses up- and down-regulated query gene sets, but unlike
CMAP it uses the quantitative differential-expression values stored in the
reference database to weight the enrichment calculation.

The method uses a bidirectional **weighted Kolmogorov-Smirnov enrichment
statistic**. Consequently, the reference database must contain quantitative
expression scores such as log2 fold changes or moderated z-scores rather than
only ranks.

## Native scores

The upstream LINCS search can report several related connectivity measures:

- `WTCS`: weighted connectivity score;
- `WTCS_Pval`: nominal p-value for WTCS;
- `WTCS_FDR`: false-discovery rate associated with the WTCS p-value;
- `NCS`: normalized connectivity score, normalized within cell and perturbation
  type;
- `NCSct`: NCS summarized across cell types;
- `Tau`: database-standardized connectivity measure, when requested.

WTCS preserves the sign of the query-reference relationship. NCS improves
comparability across cellular and perturbation contexts, while Tau expresses the
relative extremeness of connectivity against a database-specific reference
distribution.

## Main considerations

LINCS provides much broader perturbation coverage than CMAP, especially with the
LINCS2 database. At the same time, results can vary substantially across cell
types and perturbation contexts. A strong score in one reference cell line
should not automatically be interpreted as evidence of activity in the
disease-relevant tissue.

# gCMAP search

## How the method works

gCMAP differs structurally from the CMAP method.

CMAP uses **up/down query gene sets** against **ranked reference profiles**.
gCMAP uses the opposite orientation: the query is a ranked or quantitative
gene-expression profile, whereas each reference signature is represented as a
signed gene set.

For a quantitative reference database, the signed reference gene sets are
created by applying the `higher` and `lower` thresholds to each reference
profile. The numeric query values can be supplied directly and are internally
converted to ranks when required.

## Native scores

The principal gCMAP result field is:

- `effect`: scaled bidirectional enrichment effect, with positive and negative
  signs representing concordant and discordant relationships, respectively.

Other useful fields include:

- `nSet`: size of the reference gene set after thresholding;
- `nFound`: number of reference-set genes found in the query profile;
- `signed`: whether the reference sets retain separate positive and negative
  directions.

## Main considerations

The `higher` and `lower` thresholds determine which genes from each reference
profile become members of the signed gene sets. They therefore affect the
effective size and information content of the reference signatures and should
be interpreted as part of the method definition rather than as simple
computational settings.

# Correlation-based search

## How the method works

Correlation-based search directly compares a quantitative query profile with
quantitative reference profiles over their shared genes. The upstream
`signatureSearch` implementation supports Spearman, Pearson, and Kendall
correlation.

DrugSigNet commonly uses a one-column signed expression matrix as the query,
with Entrez gene identifiers stored as row names. Correlation can be calculated
over all shared genes represented by the query or over a selected subset.

A positive correlation indicates concordant profiles, whereas a negative
correlation indicates discordant profiles.

## Native score

The principal result field is:

- `cor_score`: the correlation coefficient produced by the selected correlation
  method.

## Main considerations

Correlation preserves quantitative information that is discarded by strict
up/down gene-set searches. However, query and reference values should represent
biologically comparable expression signals, and the same gene identifiers must
be used on both sides of the comparison.

Spearman correlation is rank-based and is less dependent on the absolute scale
of the expression statistic than Pearson correlation. Regardless of the chosen
coefficient, results remain sensitive to gene coverage, preprocessing, biological
context, and technical effects in the input profile.

# Method comparison

| Method | Query used in search | Reference representation | Main native score | Main methodological consideration |
|---|---|---|---|---|
| CMAP | Up/down gene sets | Ranked expression profile | `raw_score`, `scaled_score` | Query thresholding determines information retained |
| LINCS | Up/down gene sets | Quantitative expression scores | `WTCS` (plus NCS/Tau derivatives) | Weighted enrichment is context dependent |
| gCMAP | Ranked/quantitative profile | Threshold-derived signed gene sets | `effect` | `higher`/`lower` thresholds define reference sets |
| Correlation | Quantitative profile | Quantitative profile | `cor_score` | Query and reference must share comparable quantitative information |

Rather than choosing one method solely on theoretical preference, DrugSigNet is
designed to compare complementary search formulations and integrate their
rankings downstream. Agreement across methods can strengthen confidence, while
method-specific results can reveal sensitivity to the representation used for
the query or reference signature.

# Practical recommendations

- Use consistent Entrez identifiers and inspect how many query genes map to the
  selected reference database.
- Inspect and retain each method object's `@parameters` slot together with the
  results so that the exact query and search settings are documented.
- Define the disease signature using biologically and statistically justified
  criteria before running signature search.
- Interpret score **sign** separately from score **magnitude**; a strong
  relationship may be concordant or discordant.
- Compare results across both CMAP and LINCS2 because their coverage and
  perturbation contexts differ substantially.
- Treat cell-line and perturbation context as an important source of biological
  heterogeneity rather than assuming that every reference signature is equally
  relevant to the disease.
- Use consensus ranking to identify candidates supported across complementary
  methods, while retaining method-specific results for sensitivity analysis.
- Treat prioritized drugs as computational hypotheses requiring biological
  interpretation and, where appropriate, experimental validation.

# Individual method calls

The examples below demonstrate the four DrugSigNet method wrappers. The complete
multi-method orchestration, filtering, rank aggregation, annotation, and
visualization workflow is covered in `signature-workflow.Rmd`.

## Prepare query representations

```{r signature-example-input}
data("disease_signature", package = "DrugSigNet")

upset <- as.character(
  disease_signature$Entrez[disease_signature$FC > 0]
)
downset <- as.character(
  disease_signature$Entrez[disease_signature$FC < 0]
)

signature_matrix <- as.matrix(
  disease_signature[, "FC", drop = FALSE]
)
rownames(signature_matrix) <- as.character(disease_signature$Entrez)
colnames(signature_matrix) <- "FC"
```

For this compact example, positive and negative fold changes are used directly
to illustrate the two query representations. In a scientific analysis, gene
selection should follow the statistical and biological criteria appropriate to
the input study.

## CMAP

```{r run-cmap-method}
cmap_result <- cmap_method(
  upset = upset,
  downset = downset,
  ref_db = "cmap",
  chunk_size = 5000
)

head(cmap_result@result)
```

The same method can be evaluated against LINCS2 by setting
`ref_db = "lincs2"`.

## LINCS

```{r run-lincs-method}
lincs_result <- lincs_method(
  upset = upset,
  downset = downset,
  ref_db = "lincs2",
  chunk_size = 5000
)

head(lincs_result@result)
```

The LINCS algorithm can also be applied to the CMAP quantitative reference by
setting `ref_db = "cmap"`.

## gCMAP

The `higher` and `lower` arguments determine how quantitative reference
signatures are converted to signed reference gene sets.


```{r run-gcmap-method}
gcmap_result <- gcmap_method(
  signature_matrix = signature_matrix,
  ref_db = "cmap",
  higher = min(signature_matrix[signature_matrix > 0]),
  lower = max(signature_matrix[signature_matrix < 0]),
  chunk_size = 5000
)
head(gcmap_result@result)
```


These thresholds are shown for illustration. They retain positive and negative
directions in this example but should not be treated as universally optimal
settings.

## Correlation

```{r run-correlation-method}
correlation_result <- correlation_method(
  signature_matrix = signature_matrix,
  ref_db = "cmap",
  method = "spearman",
  chunk_size = 5000
)

head(correlation_result@result)
```

Correlation can likewise be run against LINCS2 by changing `ref_db` to
`"lincs2"`.

## Using a frozen local reference

For reproducible runs against a fixed HDF5 database, replace the named
`ref_db` value with the local path returned by `load_signature_refdb()`:

```{r frozen-reference-example}
# Example pattern:
# frozen_ref <- load_signature_refdb(...)
#
# cmap_result <- cmap_method(
#   upset = upset,
#   downset = downset,
#   ref_db = frozen_ref,
#   chunk_size = 5000
# )
```

See `signature-workflow.Rmd` for the full pipeline-level reference-database
workflow.

# Common pitfalls

## Too few mapped genes

Very small or poorly mapped signatures can produce unstable enrichment or
correlation results. Inspect gene overlap with the selected reference database
before interpreting rankings.

## Arbitrary query thresholds

CMAP and LINCS depend on the up/down query sets, while gCMAP depends on
thresholds used to construct reference gene sets. Arbitrary cutoffs can change
the information supplied to the search and should be justified or evaluated in
sensitivity analyses.

## Strong technical effects

Batch effects, inappropriate normalization, or compositional differences can
dominate the disease signature and consequently the connectivity search.
Quality control and suitable differential-expression analysis should be
completed before DrugSigNet is applied.

## Mixing score semantics

The native scores are not numerically interchangeable across methods:
`scaled_score`, `WTCS`, `effect`, and `cor_score` arise from different
statistical formulations. Compare ranks or appropriately harmonized downstream
results rather than treating equal numeric values from different methods as
equivalent evidence.

## Overinterpreting reference context

CMAP and LINCS2 signatures were generated in specific cellular and perturbation
contexts. A highly ranked compound does not imply that the same transcriptional
response will occur in the disease-relevant cell type, dose, exposure time, or
organism.

# Summary

This vignette described the relevant DrugSigNet concepts, inputs, outputs, and interpretation principles. Preserve the input data, function arguments, package version, and returned S4 objects when reporting or reproducing an analysis.

# Next steps

| Goal | Recommended vignette |
|---|---|
| Choose between signature, network, and integrated analysis | `choosing-analysis-strategy.Rmd` |
| Run the complete signature workflow | `signature-workflow.Rmd` |
| Understand consensus ranking | `rank-aggregation.Rmd` |
| Annotate prioritized compounds | `drug-annotation.Rmd` |
| Visualize signature and consensus results | `visualization.Rmd` |

# Session information

Record the R and package versions used to build this vignette.

```{r session-info, eval=TRUE}
sessionInfo()
```
