---
title: "Signature workflow with DrugSigNet"
author: "DrugSigNet authors"
date: "Last update: `r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{08. Signature workflow with DrugSigNet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

This vignette shows how to run and inspect the complete signature-based
DrugSigNet workflow.

The emphasis is on how the pipeline converts method-specific signature-search
results into standardized rank scores and then combines evidence across
reference databases and signature-search methods. For algorithm-specific
details, see `signature-methods.Rmd`; for the aggregation algorithms
themselves, see `rank-aggregation.Rmd`.

# Workflow overview

`drugSignaturePipeline()` performs the following steps:

1. prepares up-regulated, down-regulated, and quantitative representations of
   the disease signature;
2. runs CMAP, LINCS, gCMAP, and correlation searches against CMAP and LINCS2
   reference databases;
3. converts each raw search result to a standardized method-specific
   `rank_score`;
4. harmonizes perturbation names and the drug universe across methods;
5. aggregates the CMAP- and LINCS2-reference results within each signature
   method using CRank, Dowdall, and RRA;
6. aggregates those method-family consensus rankings across CMAP, LINCS, gCMAP,
   and correlation; and
7. returns method-specific and harmonized signature rankings for downstream
   annotation and visualization.

The eight signature-search results are:

```text
CMAP_CMAP
LINCS_CMAP
gCMAP_CMAP
Correlation_CMAP
CMAP_LINCS2
LINCS_LINCS2
gCMAP_LINCS2
Correlation_LINCS2
```

The first part of each name identifies the search method and the second part
identifies the reference database.

# Prepare the input signature

The signature workflow requires a data frame with exactly two columns:

- `Entrez`: Entrez gene identifiers;
- `FC`: signed fold-change values, typically log2 fold changes.

```{r prepare-signature}
data("disease_signature", package = "DrugSigNet")

signature_input <- disease_signature

head(signature_input)
```

Positive and negative fold changes are used to construct the up- and
down-regulated query sets required by CMAP and LINCS. The same signed profile is
converted internally to a one-column numeric matrix for gCMAP and correlation.

When the same Entrez identifier occurs more than once, the pipeline retains the
largest positive fold change for the up-regulated set and the largest absolute
negative fold change for the down-regulated set. Genes occurring in both sets
are removed before the quantitative query matrix is constructed.

# Run the signature pipeline

For an initial analysis, annotation and visualization can be disabled so that
the signature-search and ranking stages can be inspected directly.

```{r signature-run}
sig_res <- drugSignaturePipeline(
  signature_input = signature_input,
  ties_method = "max",
  prior = 0.093,
  num_bin = 200,
  n_workers = 2,
  top_k = 100,
  run_drug_annotation = FALSE,
  run_visualization = FALSE
)
```

The same signature workflow can also be called through:

```{r signature-unified-run}
sig_res_unified <- drugRepurposingPipeline(
  signature_input = signature_input,
  mode = "signature",
  ties_method = "max",
  top_k = 100,
  run_drug_annotation = FALSE,
  run_visualization = FALSE
)
```

# Important arguments

| Argument | Meaning |
|---|---|
| `signature_input` | Disease signature containing `Entrez` and `FC` |
| `trend` | Optional perturbation-direction filter, for example `"up"` or `"down"`; `NULL` retains both |
| `drug_name_synonym` | Optional synonym table used to harmonize perturbation/drug names |
| `ties_method` | Tie-handling rule used during method-specific scoring and rank aggregation |
| `prior` | CRank prior; default `0.093` |
| `num_bin` | Number of CRank rank intervals; default `200` |
| `n_workers` | Number of workers used to run signature searches |
| `chunk_size` | Number of reference signatures processed per chunk; default `5000` |
| `signature_refdb_mode` | Use `"default"`, `"frozen"`, or `"frozen_force"` reference resources |
| `padj` | Optional adjusted-p-value threshold passed to the gCMAP CMAP-reference search |
| `top_k` | Top-rank threshold used for downstream annotation/visualization; it does not truncate the core ranking tables |
| `run_drug_annotation` | Run downstream drug annotation and target enrichment |
| `run_visualization` | Generate the visualization section |

The `trend` option filters perturbation direction before method-specific rank
scores are calculated. Direction should therefore be selected according to the
biological question rather than assuming a single preferred direction.

# Inspect the result structure

`drugSignaturePipeline()` returns a `DrugSearchingPipeline` object.

```{r inspect-signature-structure}
slotNames(sig_res)

names(sig_res@DrugSearching)
names(sig_res@RankAggregation)
```

The two main signature-analysis sections are:

| Section | Contents |
|---|---|
| `@DrugSearching[["Raw"]]` | Raw method/reference search objects |
| `@DrugSearching[["Processed"]]` | Standardized method-specific drug rankings containing `rank_score` |
| `@RankAggregation[["Pairwise"]]` | Consensus across CMAP and LINCS2 references within each method family |
| `@RankAggregation[["Harmonized_methods"]]` | Consensus across the four signature-method families |
| `@RankAggregation[["Signature_Harmonized"]]` | Final analysis-ready signature ranking table |

# Raw signature-search results

Inspect the available raw searches with:

```{r inspect-raw-signature}
names(sig_res@DrugSearching[["Raw"]])

head(
  sig_res@DrugSearching[["Raw"]][["CMAP_CMAP"]]@result
)
```

Each raw entry is a `SignatureBased` object containing the native
method-specific output. Its `@parameters` slot retains the query and search
settings; see `signature-methods.Rmd` for details.

Raw native scores differ by method:

| Method | Main native quantity used for DrugSigNet ranking |
|---|---|
| CMAP | `raw_score` |
| LINCS | `WTCS` |
| gCMAP | `effect` |
| Correlation | `cor_score` |

Because these native values have different definitions and scales, they are not
combined directly.

# Method-specific rank scores

DrugSigNet converts each raw result into a common processed representation.

```{r inspect-processed-signature}
names(sig_res@DrugSearching[["Processed"]])

head(
  sig_res@DrugSearching[["Processed"]][["CMAP_CMAP"]]
)
```

The shared processed schema includes `perturbation`, `rank_score`, the
method-specific native score, `trend`, `scaled_score`, and available
perturbation metadata.

## How `rank_score` is calculated

For each method, DrugSigNet first retains the strongest absolute
drug-signature association for each perturbagen. Cell information is collapsed
when the same perturbagen is represented in multiple cells.

Qualifying perturbagens are then ranked by the **absolute magnitude** of the
method-specific score:

\[
\mathrm{rank\_score} = \frac{1}{\mathrm{rank}}.
\]

Thus, among qualifying perturbagens:

```text
rank 1  -> rank_score 1
rank 2  -> rank_score 1/2
rank 3  -> rank_score 1/3
...
```

**Larger `rank_score` values indicate stronger method-specific priority.**
Perturbagens that do not pass the method-specific criterion receive
`rank_score = 0`.

The sign/direction is retained separately in `trend`; ranking by absolute
magnitude does not discard the original direction from the processed output.

The significance or inclusion rule differs by method:

| Method | Ranking quantity | Criterion for non-zero `rank_score` | `scaled_score` |
|---|---|---|---|
| CMAP | `abs(raw_score)` | Absolute raw score passes the 95% KS critical-value cutoff | CMAP scaled score returned by the search |
| LINCS | `abs(WTCS)` | BY-adjusted `WTCS_Pval <= 0.05` | Min-max scaled `abs(WTCS)` |
| gCMAP | `abs(effect)` | Effect exceeds the cutoff derived from the corresponding significant CMAP results | `effect` |
| Correlation | `abs(cor_score)` | Correlation p-value `<= 0.05` | Min-max scaled `abs(cor_score)` |

For gCMAP, the corresponding CMAP result is used to derive the effect cutoff.
For correlation, the corresponding CMAP query size is used when calculating the
correlation significance test.

The `ties_method` argument controls ties during these reciprocal-rank
calculations.

# Harmonizing the processed drug universe

The processed method tables do not necessarily contain exactly the same
perturbagens. Before aggregation, DrugSigNet harmonizes available perturbation
metadata across methods and adds missing method/drug combinations with
`rank_score = 0`.

This creates a common representation in which zero means that a drug has no
qualifying evidence from that method, rather than forcing raw scores from
different algorithms onto a common numerical scale.

# Pairwise aggregation across reference databases

The first aggregation stage combines the two reference-database results within
each search-method family.

| Pairwise family | Inputs |
|---|---|
| CMAP | `CMAP_CMAP` + `CMAP_LINCS2` |
| LINCS | `LINCS_CMAP` + `LINCS_LINCS2` |
| gCMAP | `gCMAP_CMAP` + `gCMAP_LINCS2` |
| Correlation | `Correlation_CMAP` + `Correlation_LINCS2` |

Each pair is aggregated independently with CRank, Dowdall, and RRA.

```{r inspect-pairwise-aggregation}
names(sig_res@RankAggregation[["Pairwise"]])

names(
  sig_res@RankAggregation[["Pairwise"]][["CMAP"]]
)

sig_res@RankAggregation[["Pairwise"]][["CMAP"]][["CRank"]]@result
sig_res@RankAggregation[["Pairwise"]][["CMAP"]][["Dowdall"]]@result
sig_res@RankAggregation[["Pairwise"]][["CMAP"]][["RRA"]]@result
```

At this stage the inputs are processed `rank_score` values, where **larger is
better**. DrugSigNet therefore applies the aggregation direction appropriate to
that input:

```text
CRank:   reverse = FALSE
Dowdall: reverse = TRUE
RRA:     reverse = TRUE
```

The resulting pairwise consensus columns are named by method family, for
example `CMAP_CRank`, `CMAP_Dowdall`, and `CMAP_RRA`.

# Harmonizing across signature methods

The second aggregation stage combines the pairwise consensus ranks from CMAP,
LINCS, gCMAP, and correlation.

Importantly, this stage combines **ranks**, not the original native scores or
processed reciprocal-rank scores. Smaller pairwise consensus ranks are better,
so the aggregation direction changes accordingly:

```text
CRank:   reverse = TRUE
Dowdall: reverse = FALSE
RRA:     reverse = FALSE
```

The three harmonized aggregation objects are stored in:

```{r inspect-harmonized-methods}
names(
  sig_res@RankAggregation[["Harmonized_methods"]]
)

sig_res@RankAggregation[["Harmonized_methods"]][["CRank"]]@result
sig_res@RankAggregation[["Harmonized_methods"]][["Dowdall"]]@result
sig_res@RankAggregation[["Harmonized_methods"]][["RRA"]]@result
```

This two-stage design separates two sources of variability:

1. **reference-database variability** is handled within each signature method;
2. **methodological variability** is handled across CMAP, LINCS, gCMAP, and
   correlation.

# Final signature ranking table

The final table is:

```{r inspect-signature-harmonized}
signature_rankings <-
  sig_res@RankAggregation[["Signature_Harmonized"]]

head(signature_rankings)
colnames(signature_rankings)
```

`Signature_Harmonized` contains:

- one rank column for each processed method/reference combination; and
- the final signature consensus ranks:
  - `Signature_CRank`;
  - `Signature_Dowdall`;
  - `Signature_RRA`.

Before these method-specific columns are added to the final table, the processed
`rank_score` values are converted from larger-is-better scores to
smaller-is-better ranks. The current implementation uses `"max"` tie handling
for this final method-column conversion.

The final consensus columns are already ranks, so **smaller values indicate
higher signature-based priority**.

For example:

```{r top-signature-consensus}
signature_rankings |>
  dplyr::arrange(Signature_CRank) |>
  dplyr::select(
    Drug,
    Signature_CRank,
    Signature_Dowdall,
    Signature_RRA
  ) |>
  head(20)
```

Agreement across the three consensus formulations can be inspected without
assuming that one aggregation method is universally preferable.

# Reference-database reproducibility

By default, DrugSigNet uses the reference resources resolved through
`signatureSearch`. Frozen Synapse-backed CMAP and LINCS2 HDF5 resources can be
used when the analysis needs to be repeated against fixed reference files.

```{r frozen-signature-reference}
sig_frozen <- drugSignaturePipeline(
  signature_input = signature_input,
  signature_refdb_mode = "frozen",
  auth_token = Sys.getenv("SYNAPSE_AUTH_TOKEN"),
  ties_method = "max",
  prior = 0.093,
  num_bin = 200,
  n_workers = 2,
  run_drug_annotation = FALSE,
  run_visualization = FALSE
)
```

See `signature-methods.Rmd` for reference-database provenance and individual
method parameters.

# Downstream annotation and visualization

Annotation and visualization can be enabled directly in the signature pipeline
with:

```text
run_drug_annotation = TRUE
run_visualization = TRUE
```

When annotation is enabled, DrugSigNet preferentially selects the union of the
top `top_k` drugs across `Signature_CRank`, `Signature_Dowdall`, and
`Signature_RRA` for downstream target-based summaries.

Detailed annotation and plotting interfaces are documented in
`drug-annotation.Rmd` and `visualization.Rmd`.

# Exporting and reporting results

Completed signature pipeline results can be exported with
`write_pipeline_results()` or summarized in an HTML/PDF report with
`write_report()`. See `complete-case-study.Rmd` for complete examples.

# Interpretation checklist

- Are the disease-signature identifiers compatible with the reference
  databases?
- Do the eight processed searches provide broadly consistent or complementary
  evidence?
- Are high-ranking drugs supported across both CMAP and LINCS2 references?
- Do the pairwise CMAP, LINCS, gCMAP, and correlation consensus rankings agree?
- Do `Signature_CRank`, `Signature_Dowdall`, and `Signature_RRA` identify a
  stable set of high-priority candidates?
- Is the retained connectivity direction biologically plausible for the disease
  state and analysis objective?

# Troubleshooting

## All signature methods fail

Check reference-database availability and Entrez identifier compatibility. If
the error mentions ExperimentHub, Synapse, or a frozen HDF5 resource, check
network access, authentication, and the local cache.

## Too few drugs receive non-zero rank scores

Inspect the method-specific filtering criteria, gene mapping, optional
`trend`, and drug synonym mapping. A drug may be present in the common processed
universe while receiving `rank_score = 0` for a method that does not provide
qualifying evidence.

## Pairwise aggregation is missing for one method family

Pairwise aggregation requires both the CMAP- and LINCS2-reference processed
results for that family. If either search is unavailable, that family is
skipped during pairwise aggregation.

## Results differ across reference databases

This is expected because CMAP and LINCS2 differ in platform, scale, compounds,
cells, and perturbation conditions. The pairwise aggregation stage is designed
to summarize evidence across those reference resources before the
method-family consensus is calculated.

# Summary

This vignette described the relevant DrugSigNet concepts, inputs, outputs, and interpretation principles. Preserve the input data, function arguments, package version, and returned S4 objects when reporting or reproducing an analysis.

# Next steps

| Goal | Recommended vignette |
|---|---|
| Understand individual signature methods | `signature-methods.Rmd` |
| Understand CRank, Dowdall, and RRA | `rank-aggregation.Rmd` |
| Annotate prioritized drugs | `drug-annotation.Rmd` |
| Visualize signature rankings | `visualization.Rmd` |
| Run the network workflow | `network-workflow.Rmd` |
| Integrate signature and network evidence | `complete-case-study.Rmd` |

# Session information

Record the R and package versions used to build this vignette.

```{r session-info, eval=TRUE}
sessionInfo()
```
