---
title: "Visualization with DrugSigNet"
author: "DrugSigNet authors"
date: "Last update: `r format(Sys.Date(), '%d %B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{07. Visualization with DrugSigNet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)
```

```{r load-package, include=FALSE}
library(DrugSigNet)
```

# Introduction

DrugSigNet provides standalone plotting functions for ranked drug lists,
annotation summaries, target-set enrichment, and drug-drug similarity.

This vignette summarizes the current visualization API. Complete signature,
network, and integrated analyses are covered in the workflow vignettes.

# Visualization functions

| Function | Purpose | Return |
|---|---|---|
| `plot_drug_hierarchy()` | Hierarchical drug annotations | Plotly sunburst or ggplot alluvial plot |
| `plot_drug_indications()` | Indication frequencies | ggplot bar plot or recorded word cloud |
| `plot_drug_signature_overlay()` | Drug-target, disease, and signature network overlay | List containing the plot and evidence tables |
| `plot_drug_similarity()` | Drug-drug similarity | ggplot heatmap or invisible `hclust` |
| `plot_drug_status()` | Drug and condition-specific status | ggplot |
| `plot_enriched_terms()` | Enrichment results | ggplot |
| `plot_rank_agreement()` | Pairwise rank correlations | List containing matrices and `plot` |
| `plot_rank_distribution()` | Pairwise rank scatter plots | ggplot |
| `plot_top_k_hits()` | Annotated hits among top-ranked drugs | ggplot |
| `plot_top_k_overlap()` | Top-k overlap across methods | ggplot |
| `plot_top_k_overlap_comparison()` | Top-k overlap between two ranking results | ggplot |
| `plot_all()` | Run multiple plotting helpers | List containing `plots` and `errors` |

Most functions accept analysis-ready data frames directly; a complete pipeline
object is not required.

# Drug annotation plots

## Drug hierarchy

`plot_drug_hierarchy()` displays hierarchical annotation columns as a sunburst
or alluvial plot. At least two hierarchy columns are required.

```{r plot-drug-hierarchy}
hierarchy_df <- data.frame(
  ATC = rep(
    c("Nervous system", "Antineoplastic", "Cardiovascular", "Metabolic"),
    each = 4
  ),
  MoA = rep(
    c(
      "Neurotransmitter modulator", "Kinase inhibitor",
      "Ion-channel modulator", "Metabolic enzyme inhibitor"
    ),
    each = 4
  ),
  Drug = paste0("drug_", sprintf("%02d", 1:16)),
  N = c(18, 14, 11, 8, 17, 15, 10, 7, 16, 13, 9, 6, 14, 12, 8, 5)
)

p1 <- plot_drug_hierarchy(
  data_df = hierarchy_df,
  hierarchy_cols = c("ATC", "MoA", "Drug"),
  value_col = "N",
  plot_type = "sunburst"
)

p2 <- plot_drug_hierarchy(
  data_df = hierarchy_df,
  hierarchy_cols = c("ATC", "MoA", "Drug"),
  value_col = "N",
  plot_type = "alluvial"
)

p1
p2
```

Use `layer_order = "inner_to_outer"` to reverse the hierarchy and `na_label`
to control the label assigned to missing categories. The sunburst (`p1`) is
interactive, whereas the alluvial view (`p2`) provides a static alternative.

## Drug indications

`plot_drug_indications()` requires `Drug` and `indication` columns. Multiple
indications separated by `|` or `;` are split before counting.

```{r plot-drug-indications}
indication_df <- data.frame(
  Drug = paste0("drug_", sprintf("%02d", 1:24)),
  indication = rep(
    c(
      "Alzheimer disease|Cognitive impairment",
      "Cancer|Solid tumor",
      "Diabetes;Metabolic disease",
      "Cardiovascular disease|Hypertension",
      "Inflammation|Autoimmune disease",
      "Asthma|Respiratory disease"
    ),
    each = 4
  ),
  stringsAsFactors = FALSE
)

p <- plot_drug_indications(
  data_df = indication_df,
  plottype = "barplot",
  top_terms = 10
)

p
```

A word cloud can be generated with:

```{r plot-drug-indications-wordcloud}
wc <- plot_drug_indications(
  data_df = indication_df,
  plottype = "wordcloud",
  max_words = 50
)

wc
```

## Drug status

`plot_drug_status()` creates a nested donut chart from `drug`,
`highest_status`, and `condition_status`.

```{r plot-drug-status}
status_df <- data.frame(
  drug = paste0("drug_", sprintf("%02d", 1:20)),
  highest_status = rep(
    c("Approved", "Phase 3", "Phase 2", "Phase 1", "Preclinical"),
    each = 4
  ),
  condition_status = rep(
    c("Approved", "Clinical", "Clinical", "Preclinical"),
    length.out = 20
  )
)

plot_drug_status(status_df)
```

The inner ring represents `condition_status` and the outer ring represents
`highest_status`. Categories can be removed with `exclude_condition_status`.

# Drug similarity

`plot_drug_similarity()` accepts a square numeric similarity matrix and can
produce either a heatmap or a dendrogram.

```{r plot-drug-similarity}
set.seed(42)
drug_features <- matrix(rnorm(15 * 8), nrow = 15)
sim_mat <- stats::cor(t(drug_features))
sim_mat <- (sim_mat + 1) / 2
diag(sim_mat) <- 1
drug_labels <- paste0("drug_", sprintf("%02d", 1:15))
dimnames(sim_mat) <- list(drug_labels, drug_labels)

plot_drug_similarity(
  sim_mat,
  plot_type = "heatmap"
)

plot_drug_similarity(
  sim_mat,
  plot_type = "dendrogram"
)
```

Dendrogram mode clusters drugs using `1 - similarity`. Heatmaps can be shown as
`"triangle"` or `"full"` using `heatmap_type`.

# Drug signature network overlay

`plot_drug_signature_overlay()` combines a drug-centered interaction subgraph
with disease seed genes and genes from a drug signature. The example below
uses a small, illustrative perhexiline network so that it can be built without
downloading an external interaction network or signature database. The edges
are for demonstration only and should not be interpreted as a curated
perhexiline mechanism.

```{r plot-drug-signature-overlay, fig.width=9, fig.height=7}
perhexiline_subgraph <- list(
  drug = "perhexiline",
  subgraph = data.frame(
    from = c(
      "perhexiline", "perhexiline", "CPT1A", "CPT2",
      "CPT1A", "NDUFS1"
    ),
    to = c("CPT1A", "CPT2", "APOE", "PPARG", "NDUFS1", "HIF1A"),
    type = c("DT", "DT", "PPI", "PPI", "PPI", "PPI")
  )
)

perhexiline_disease_seeds <- data.frame(
  gene = c("CPT2", "APOE"),
  direction = c("down", "up")
)

perhexiline_signature <- data.frame(
  gene = c("CPT1A", "HIF1A"),
  direction = c("down", "up")
)

perhexiline_gene_map <- data.frame(
  gene = c("CPT1A", "CPT2", "APOE", "PPARG", "NDUFS1", "HIF1A"),
  symbol = c("CPT1A", "CPT2", "APOE", "PPARG", "NDUFS1", "HIF1A")
)

perhexiline_overlay <- plot_drug_signature_overlay(
  Drug = "perhexiline",
  subgraph_result = perhexiline_subgraph,
  disease_seed_genes = perhexiline_disease_seeds,
  drug_signature_genes = perhexiline_signature,
  symbol_map = perhexiline_gene_map,
  include_all_drug_targets = TRUE,
  seed = 42
)

perhexiline_overlay$plot
```

The larger drug node is connected to its direct targets. Node color separates
disease seeds, drug-signature genes, and other targets, while emphasized node
outlines identify direct drug targets. The returned object also contains
`disease_seed_evidence`, `drug_signature_evidence`, and `drug_sig_path_info`
tables, which can be inspected to understand why each gene was included.

# Enrichment plots

`plot_enriched_terms()` visualizes enrichment results such as those returned by
`TSEA()`. The input must contain `Term`, `ont`, `Adjusted.P.value`, and
`Overlap`, where `Overlap` is formatted as `"x/y"`.

```{r plot-enriched-terms}
enrich_df <- data.frame(
  Term = c(
    "apoptotic process (GO:0006915)",
    "cell cycle (GO:0007049)",
    "inflammatory response (GO:0006954)",
    "oxidative stress response (GO:0006979)",
    "synaptic signaling (GO:0099536)",
    "protein phosphorylation (GO:0006468)",
    "kinase activity (GO:0016301)",
    "receptor binding (GO:0005102)",
    "transcription regulator activity (GO:0140110)",
    "protein complex binding (GO:0032403)",
    "mitochondrial matrix (GO:0005759)",
    "synaptic membrane (GO:0097060)"
  ),
  ont = rep(c("BP", "MF", "CC"), each = 4),
  Adjusted.P.value = c(
    0.0002, 0.0008, 0.002, 0.006,
    0.0005, 0.001, 0.004, 0.012,
    0.0009, 0.003, 0.009, 0.025
  ),
  Overlap = c(
    "18/220", "15/190", "13/170", "11/150",
    "16/210", "14/180", "12/160", "9/130",
    "15/200", "12/165", "10/145", "8/120"
  )
)

plot_enriched_terms(
  enrich_df,
  plottype = "dotplot"
)

plot_enriched_terms(
  enrich_df,
  plottype = "barplot"
)
```

Use `show_category` to control the number of terms displayed per ontology.

# Ranking comparison plots

The ranking functions below expect numeric rank columns, where **smaller values
represent higher priority**.

## Rank agreement

`plot_rank_agreement()` calculates pairwise rank correlations and returns both
the plot and the underlying correlation statistics.

```{r plot-rank-agreement}
set.seed(42)
n_drugs <- 100
base_rank <- seq_len(n_drugs)
rank_df <- data.frame(
  Drug = paste0("drug_", sprintf("%03d", seq_len(n_drugs))),
  CMAP = base_rank,
  LINCS = rank(base_rank + rnorm(n_drugs, sd = 12)),
  gCMAP = rank(base_rank + rnorm(n_drugs, sd = 20)),
  Correlation = rank(rev(base_rank) + rnorm(n_drugs, sd = 18))
)

agreement_res <- plot_rank_agreement(
  drug_ranks_df = rank_df,
  cor_method = "spearman",
  top_k = 100
)

agreement_res$plot
agreement_res$cor_matrix
agreement_res$p_matrix
```

Supported correlation methods are `"spearman"`, `"pearson"`, and `"kendall"`.
With `cluster = TRUE`, methods are reordered by their correlation structure.

## Rank distributions

`plot_rank_distribution()` creates pairwise scatter plots across selected
ranking methods.

```{r plot-rank-distribution}
rank_distribution_df <- rank_df

plot_rank_distribution(rank_distribution_df)

plot_rank_distribution(
  drug_ranks_df = rank_distribution_df,
  methods = c("CMAP", "LINCS", "gCMAP"),
  top_k = 100
)
```

When `methods = NULL`, all numeric columns are used.

# Top-k plots

## Top-k hits

`plot_top_k_hits()` compares status-annotated hits among top-ranked drugs.
Status can be supplied in a separate data frame or stored in the ranking table.

```{r plot-top-k-hits}
top_k_rank_df <- data.frame(
  Drug = c(
    "daratumumab|felzartamab|isatuximab",
    paste0("drug_", sprintf("%03d", 2:80))
  ),
  Method1 = seq_len(80),
  Method2 = rank(seq_len(80) + sin(seq_len(80) / 5) * 12),
  Method3 = rank(seq_len(80) + cos(seq_len(80) / 7) * 15)
)

hit_status_df <- data.frame(
  Drug = c(
    "daratumumab", "felzartamab", "isatuximab",
    paste0("drug_", sprintf("%03d", seq(5, 75, by = 5)))
  ),
  Status = "positive"
)

plot_top_k_hits(
  drug_ranks_df = top_k_rank_df,
  status_df = hit_status_df,
  split_grouped_drugs = FALSE,
  plottype = "Barplot"
)
```

Supported plot types are `"Barplot"`, `"dotplot"`, and `"lineplot"`.

For grouped drug labels, `split_grouped_drugs = FALSE` keeps the group as one
ranked entity while still matching individual members to status annotations.
Use `TRUE` to count grouped members separately after top-k selection.

## Top-k overlap

`plot_top_k_overlap()` compares pairwise overlap among top-ranked drug sets.

```{r plot-top-k-overlap}
overlap_rank_df <- top_k_rank_df

overlap_status_df <- data.frame(
  Drug = hit_status_df$Drug,
  Status = hit_status_df$Status
)

p_overlap <- plot_top_k_overlap(
  drug_ranks_df = overlap_rank_df,
  status_df = overlap_status_df,
  top_k = 20,
  split_grouped_drugs = FALSE
)

p_overlap
p_overlap$data
```

A drug is included when its rank is less than or equal to `top_k`. When status
information is available, cell labels can also report the number of overlapping
status-annotated drugs.

## Compare two ranking results

`plot_top_k_overlap_comparison()` compares methods from two separate ranking
tables. Drug identifiers are matched case-insensitively.

```{r plot-top-k-overlap-comparison}
set.seed(17)
comparison_drugs <- paste("Drug", sprintf("%03d", 1:100))
df_a <- data.frame(
  Drug = comparison_drugs,
  Method_A1 = seq_len(100),
  Method_A2 = rank(seq_len(100) + rnorm(100, sd = 15))
)

df_b <- data.frame(
  Drug = tolower(c(comparison_drugs[1:85], paste("Drug", sprintf("%03d", 101:115)))),
  Method_B1 = rank(seq_len(100) + rnorm(100, sd = 18)),
  Method_B2 = rank(rev(seq_len(100)) + rnorm(100, sd = 20))
)

status_data <- data.frame(
  Drug = comparison_drugs[seq(1, 100, by = 4)],
  Status = rep(c("Strong", "Moderate", "Weak", "No effect", NA), 5)
)

comparison_plot <- plot_top_k_overlap_comparison(
  df_x = df_a,
  df_y = df_b,
  top_k = 100,
  status_df = status_data
)

comparison_plot
comparison_plot$data
```

Use `methods_x` and `methods_y` to restrict or order the methods included in the
comparison.

# Generate multiple plots with `plot_all()`

`plot_all()` runs multiple plotting helpers from one named input list.

Supported keys are:

```text
drug_status
drug_indications
drug_similarity
drug_hierarchy
top_k_hits
top_k_overlap
enriched_terms
rank_agreement
rank_distribution_scatter
```

```{r plot-all}
plot_all_rank_df <- data.frame(
  Drug = rank_df$Drug,
  Status = ifelse(seq_len(nrow(rank_df)) %% 6 == 0, "positive", NA),
  CMAP = rank_df$CMAP,
  LINCS = rank_df$LINCS,
  gCMAP = rank_df$gCMAP
)

plot_all_indication_df <- data.frame(
  Drug = plot_all_rank_df$Drug[1:30],
  indication = rep(
    c(
      "Neurodegeneration|Cognitive impairment",
      "Cancer|Solid tumor",
      "Metabolic disease|Diabetes",
      "Inflammation|Autoimmune disease",
      "Cardiovascular disease|Hypertension"
    ),
    each = 6
  )
)

all_plots <- plot_all(
  plot_inputs = list(
    top_k_hits = list(
      drug_ranks_df = plot_all_rank_df,
      plottype = "Barplot"
    ),
    top_k_overlap = plot_all_rank_df,
    drug_indications = plot_all_indication_df
  ),
  file_type = "pdf",
  output_dir = "plots",
  file_prefix = "DrugSigNet_"
)

all_plots$plots
all_plots$errors
```

With `continue_on_error = TRUE`, failed plots are reported in `$errors` while
the remaining plots continue. `plot_top_k_overlap_comparison()` is a standalone
helper and is not a `plot_all()` key.

# Saving figures

Most plotting functions can save directly by supplying `file_name` and
`file_type`.

```{r save-plot-directly}
plot_enriched_terms(
  data_df = enrich_df,
  plottype = "dotplot",
  file_name = "enriched_terms",
  file_type = "pdf",
  width = 10,
  height = 7
)
```

For ggplot objects, `ggplot2::ggsave()` can also be used.

```{r save-with-ggsave}
p <- plot_rank_distribution(
  drug_ranks_df = rank_distribution_df,
  methods = c("CMAP", "LINCS", "gCMAP")
)

ggplot2::ggsave(
  filename = "rank_distribution.pdf",
  plot = p,
  width = 7,
  height = 5
)
```

# Practical recommendations

- Use consistent drug identifiers and method names across related figures.
- Confirm that rank-based inputs contain ranks rather than raw scores.
- Report `top_k` in figure captions.
- Use the same grouped-drug handling across hit and overlap analyses.
- Restrict large annotation or enrichment tables to interpretable subsets.
- Keep the input tables used to generate publication figures.

# Common pitfalls

- **Scores supplied as ranks:** top-k and rank-comparison functions assume
  smaller values are better.
- **Extra numeric metadata:** automatically detected numeric columns may be
  interpreted as ranking methods.
- **Missing annotation columns:** annotation plots require their documented input
  schemas.
- **Invalid similarity matrix:** `plot_drug_similarity()` requires a square
  numeric matrix with at least two drugs.
- **Crowded figures:** reduce the number of methods, drugs, or categories before
  plotting.

# Summary

This vignette described the relevant DrugSigNet concepts, inputs, outputs, and interpretation principles. Preserve the input data, function arguments, package version, and returned S4 objects when reporting or reproducing an analysis.

# Next steps

| Goal | Recommended vignette |
|---|---|
| Understand consensus rankings | `rank-aggregation.Rmd` |
| Prepare annotations and TSEA results | `drug-annotation.Rmd` |
| Run the complete signature workflow | `signature-workflow.Rmd` |
| Run the complete network workflow | `network-workflow.Rmd` |
| Run an integrated analysis | `complete-case-study.Rmd` |

# Session information

Record the R and package versions used to build this vignette.

```{r session-info, eval=TRUE}
sessionInfo()
```
